---
layout: single
title: "Understand the vector data type in R and classes including strings, numbers and logicals - Data Science for scientists 101"
excerpt: "This tutorial introduces vectors in R. It also introduces the differences between strings, numbers and logical or boolean values (True / False) in R."
authors: ['Data Carpentry', 'Leah Wasser']
category: [course-materials]
class-lesson: ['get-to-know-r']
permalink: /course-materials/earth-analytics/week-2/work-with-data-types-r/
nav-title: 'Vectors in R'
dateCreated: 2016-12-13
modified: '`r format(Sys.time(), "%Y-%m-%d")`'
week: 2
sidebar:
  nav:
author_profile: false
comments: true
order: 3
course: "earth-analytics"
topics:
  reproducible-science-and-programming: ['RStudio']
---
{% include toc title="In This Lesson" icon="file-text" %}


<div class='notice--success' markdown="1">

## <i class="fa fa-graduation-cap" aria-hidden="true"></i> Learning Objectives
At the end of this activity, you will be able to:

* Understand the structure of and be able to create a vector object in R.

## <i class="fa fa-check-square-o fa-2" aria-hidden="true"></i> What you need

You need R and RStudio to complete this tutorial. Also we recommend have you
have an `earth-analytics` directory setup on your computer with a `/data`
directory with it.

* [How to Setup R / R Studio](/course-materials/earth-analytics/week-1/setup-r-rstudio/)
* [Setup your working directory](/course-materials/earth-analytics/week-1/setup-working-directory/)


</div>


## Vectors and data types

A vector is the most common data structure in `R`. A vector is defined as a
group of values, which most often are either numbers or characters. You can
assign this list of values to an object or variable, just like you
can for a single value. For example we can create a vector of animal weights:

```{r weight-vector }
weight_g <- c(50, 60, 65, 82)
weight_g

```

A vector can also contain characters:

```{r, purl=FALSE}
animals <- c("mouse", "rat", "dog")
animals
```

There are many functions that allow you to inspect the content of a
vector. `length()` tells you how many elements are in a particular vector:

```{r, purl=FALSE}
length(weight_g)
length(animals)
```

## Vector data types

An important feature of a vector, is that all of the elements are the same data
type. The function `class()` shows us the class (the data type) of an object:

```{r, purl=FALSE}
class(weight_g)
class(animals)
```

The function `str()` shows us the **structure** of the object and the elements it
contains. `str()` is a really useful function when working with large and complex
objects:

```{r, purl=FALSE}
str(weight_g)
str(animals)
```

You can add elements to your vector by using the `c()` function:

```{r, purl=FALSE}

# add the number 90 to the end of the vector
weight_g <- c(weight_g, 90)

# add the number 30 to the beginning of the vector
weight_g <- c(30, weight_g)
weight_g
```

In the examples above, we saw 2 of the 6 **atomic vector** types that `R` uses:

1. `"character"` and
2. `"numeric"`.

These are the basic data tpes that all `R` objects are built
from. The other 4 are:

* `"logical"` for `TRUE` and `FALSE` (the boolean data type)
* `"integer"` for integer numbers (e.g., `2L`, the `L` indicates to R that it's an integer)
* `"complex"` to represent complex numbers with real and imaginary parts (e.g.,
  `1+4i`) and that's all we're going to say about them
* `"raw"` that we won't discuss further

## Data type vs. data structure
Vectors are one of the many **data structures** that `R` uses. Other important
ones include: lists (`list`), matrices (`matrix`), data frames (`data.frame`) and
factors (`factor`). We will look at `data.frames` when we open our `boulder_precip`
data in the next lesson!

<div class="notice--warning" markdown="1">

## <i class="fa fa-pencil-square-o" aria-hidden="true"></i> Optional challenge activity

* **Question**: What happens when we create a vector that contains both numbers
and character values? Give it a try and write down the answer.
<!-- * _Answer_: R implicitly converts them to all be the same type -->

* **Question**: What will happen in each of these examples? (hint: use `class()`
  to check the data type of your objects):

```r
num_char <- c(1, 2, 3, 'a')
num_logical <- c(1, 2, 3, '2.45')
char_logical <- c('a', 'b', 'c', frog)
tricky <- c(1, 2, 3, '4')
```
* **Question**: Why do you think it happens?
<!-- * _Answer_: Vectors can be of only one data type. R tries to convert (=coerce)
  the content of this vector to find a "common denominator". -->

* **Question**: Can you draw a diagram that represents the hierarchy of the data
  types?
<!-- * _Answer_: `logical -> numeric -> character <-- logical` -->

</div>



```{r, echo=FALSE, eval=FALSE, purl=TRUE}
## We’ve seen that atomic vectors can be of type character, numeric, integer, and
## logical. But what happens if we try to mix these types in a single
## vector?

## What will happen in each of these examples? (hint: use `class()` to
## check the data type of your object)
num_char <- c(1, 2, 3, 'a')

num_logical <- c(1, 2, 3, TRUE)

char_logical <- c('a', 'b', 'c', TRUE)

tricky <- c(1, 2, 3, '4')

## Why do you think it happens?

## Can you draw a diagram that represents the hierarchy of the data
## types?
```


## Subsetting vectors

If we want to extract one or several values from a vector, we must provide one
or several indices in square brackets. For instance:

```{r, results='show', purl=FALSE}
animals <- c("mouse", "rat", "dog", "cat")
animals[2]
animals[c(3, 2)]
```

<i fa fa-star></i>**Data Tip:** R indexes start at 1. Programming languages like
Fortran, MATLAB, and R start
counting at 1, because that's what human beings typically do. Languages in the C
family (including C++, Java, Perl, and Python) count from 0 because that's
simpler for computers to do.
{: .notice--success}

## Subset Vectors

We can subset vectors too. For instance, if you wanted to select only the
values above 50:

```{r, results='show', purl=FALSE}
weight_g > 50    # will return logicals with TRUE for the indices that meet the condition
## so we can use this to select only the values above 50
weight_g[weight_g > 50]
```

You can combine multiple tests using `&` (both conditions are true, AND) or `|`
(at least one of the conditions is true, OR):

```{r, results='show', purl=FALSE}
weight_g[weight_g < 30 | weight_g > 50]
weight_g[weight_g >= 30 & weight_g == 21]
```

When working with vectors of characters, if you are trying to combine many
conditions it can become tedious to type. The function `%in%` allows you to test
if a value is found in a vector:

```{r, results='show', purl=FALSE}
animals <- c("mouse", "rat", "dog", "cat")
animals[animals == "cat" | animals == "rat"] # returns both rat and cat
animals %in% c("rat", "cat", "dog", "duck")
animals[animals %in% c("rat", "cat", "dog", "duck")]
```

<div class="notice--warning" markdown="1">

## <i class="fa fa-pencil-square-o" aria-hidden="true"></i> Optional challenge

* Can you figure out why `"four" > "five"` returns `TRUE`?
</div>

```{r, echo=FALSE, purl=TRUE}
# * Can you figure out why `"four" > "five"` returns `TRUE`?
```

<!--
```{r, purl=FALSE}
## Answers
## * When using ">" or "<" on strings, R compares their alphabetical order. Here
##   "four" comes after "five", and therefore is "greater than" it.
```
-->
