---
layout: single
title: "Access secure data connections using the RCurl R package."
excerpt: "This lesson reviews how to use functions within the RCurl package to access data on a secure (https) server in R. "
authors: ['Carson Farmer', 'Leah Wasser', 'Max Joseph']
modified: '`r format(Sys.time(), "%Y-%m-%d")`'
category: [course-materials]
class-lesson: ['intro-APIs-r']
permalink: /course-materials/earth-analytics/week-10/access-gapminder-data-rcurl-r/
nav-title: "Get Data From Github"
week: 10
sidebar:
  nav:
author_profile: false
comments: true
order: 3
---


{% include toc title="In This Lesson" icon="file-text" %}

<div class='notice--success' markdown="1">

## <i class="fa fa-graduation-cap" aria-hidden="true"></i> Learning Objectives

After completing this tutorial, you will be able to:

* Access data from a secure website using `read.csv()`
* Be able to describe the key difference between a `.tsv` and a `.csv` file.
* Use pipes ( `%>%` ) to send data directly to ggplot to plot!

## <i class="fa fa-check-square-o fa-2" aria-hidden="true"></i> What you need

You will need a computer with internet access to complete this lesson and the
data that we already downloaded for week 6 of the course.

</div>

```{r echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning=FALSE)

```

```{r}
# load libraries
library(dplyr)
library(ggplot2)
# optional
library(RCurl)
```

## Download gapminder data with RCurl

Next, we will use functions in the RCurl package to download data from GitHub.
Note that this may work without RCurl but we will practice using RCurl in the examples
below. GitHub has secure (https) url's.

#### Gapminder Data

Let's grab a gapminder data subset from a secure URL located on
a GitHub website. The
<a href=" http://www.gapminder.org/about-gapminder/" target="_blank">Gapminder data</a>,
contain a suite of census-like metrics that describe global development. The data
are ideal to experiment with when learning about plotting and working with data
in a tool like `R`.

<a href="https://twitter.com/JennyBryan" target="_blank">@jennybryan</a>
provides an `R` package to access the
<a href="http://www.gapminder.org/data/" target="_blank">Gapminder data</a> for
teaching. However, we will instead use `RCurl` to get it from
<a href="https://github.com/jennybc/gapminder" target="_blank">Jenny Bryan's Github Page</a>
to practice using RCurl functions.

```{r message=FALSE}
# Store base url (note the secure -- https:// -- url)
file_url = "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/gapminder.tsv"
# import the data!
gap_data = read.csv(file_url)
head(gap_data)
```

Looking at our results, we notice there is a `\t` between each data element.
This is not what we would expect when we import file into R. What is going on?

### .tsv file format

Note that the data format that we are using here is `.tsv` - which stands for
Tab Separate Values. The difference between .tsv and .csv is the separator:

* `.csv` uses a COMMA (`,`) to separate individual values in each column and row of the data
* `.tsv` uses a TAB (`\t`) to separate individual values in each column / row of the data.

We can use the `read.csv()` function to read in the tsv format. However, we need 
to tell R what the separator is. In this case it's `\t`. We can account
for this separator with the `sep=` argument.

```{r}
# Use textConnection to read content of temp as tsv
gap_data <- read.csv(file_url,
                     sep="\t")
head(gap_data)
```

That looks better. 

## Get Data with getURL()

IF you have issues using read.csv(), you can try the code below which uses the 
RCurl library. In past versions of R, Windows users often had issues with 
secure URL's in R. 

```{r, eval=FALSE}
# Use textConnection to read content of temp as tsv
# NOTE: you may not need to use this code if read.csv works above!
gap_data = read.csv(textConnection(gapminder_data_url),
                     sep="\t")
head(gap_data)
```


Next, we can summarize and plot the data! Notice that when we import the data
from github, using `read.csv()`, it imports into a `data.frame` format. Given
it's a `data.frame`, we can plot the data using `ggplot()` like we are used to.

Below, we first summarize the data by median life expectancy per year per continent.
Then we create box plots - one for each continent.

<!-- #tidylife -->

```{r life-by-continent, fig.cap="GGPLOT of gapminder data - life expectance by continent" }
# summarize the data - median value by content and year
summary_life_exp <-  gap_data %>%
   group_by(continent, year) %>%
   summarise(median_life = median(lifeExp))

ggplot(summary_life_exp, aes(x=year, y=median_life, colour = continent)) +
  geom_point() +
      labs(x="Continent",
           y="Life Expectancy (years)",
          title="Gapminder Data - Life Expectancy",
          subtitle = "Downloaded from Jenny Bryan's Github Page")

```

### Piping data to ggplot()

Above, we used dplyr pipes to summarize the data that we wanted to plot. 
A cool trick that we can use is to send our data directly to ggplot from the pipe.
This is useful if we don't need to create a new data.frame object to use in 
another part of our analysis. 

```{r life-by-continent-pipes, fig.cap="GGPLOT of gapminder data - life expectance by continent piped" }
# summarize the data - median value by content and year
gap_data %>%
   group_by(continent, year) %>%
   summarise(median_life = median(lifeExp)) %>% 
   ggplot(aes(x=year, y=median_life, colour = continent)) +
      geom_point() +
      labs(x="Continent",
           y="Life Expectancy (years)",
          title="Gapminder Data - Life Expectancy",
          subtitle = "Data piped directly into GGPLOT! Plot looks the same!")

```


Below, we make a boxplot of `lifeExp` by `continent` too. Notice in this case
we are using the dplyr output above, again. Thus it made sense above to 
save our dplyr output as a new data.frame.

```{r box-plot-by-continent, fig.cap="GGPLOT of gapminder data - life expectance by continent boxplot"}
# create box plot
ggplot(summary_life_exp,
       aes(continent, median_life)) +
      geom_boxplot()+
      labs(x="Continent",
           y="Life Expentancy (years)",
          title="Gapminder Data - Life Expectancy",
          subtitle = "Downloaded from Jenny Bryan's Github Page using getURL")

```

We can also create a more advanced plot - overlaying the data points on top of
our box plot. See the <a href="http://docs.ggplot2.org" target="_blank"> ggplot documentation</a> to learn more advanced `ggplot()` plotting approaches.

```{r box-plot-point-outliers, fig.cap="GGPLOT of gapminder data - life expectance by continent with jitter and outliers."}
ggplot(gap_data, aes(x=continent, y=lifeExp)) +
  geom_boxplot(outlier.colour="hotpink") +
      labs(x = "Continent",
           y = "Life Expectancy (years)",
           title = "Gapminder Data - Life Expectancy",
           subtitle = "Downloaded from Jenny Bryan's Github Page using getURL")
```

Or create a box plot with the data points overlaid on top.

```{r box-plot-point-jitter, fig.cap="GGPLOT of gapminder data - life expectance by continent with jitter and outliers."}
ggplot(gap_data, aes(x=continent, y=lifeExp)) +
  geom_boxplot() +
  geom_jitter(position=position_jitter(width=0.1, height=0), alpha=0.25)+
      labs(x = "Continent",
           y = "Life Expectancy (years)",
           title = "Gapminder Data - Life Expectancy",
           subtitle = "Data points overlaid on top of the box plot.")
```

If you are going to grab many of `csv` files from secure `urls`, you
might want to use a securl connection via RCurl. The function below will take a 
URL and return a csv.

```{r}
read_secure_csv_file <- function(url, the_sep=",") {
  url <- getURL(url)
  the_data <- read.csv(textConnection(url), sep=the_sep)
  return(the_data)
}

```

<i class="fa fa-lightbulb-o" aria-hidden="true"></i> **Data Tip:** The web
changes constantly! Data available via a *particular* API at a *particular* point
in time may not be available indefinitely. Consider documenting workflows carefully.
{: .notice--success}
