---
layout: single
title: "dNBR with MODIS in R"
excerpt: ". "
authors: ['Megan Cattau', 'Leah Wasser']
modified: '`r format(Sys.time(), "%Y-%m-%d")`'
category: [course-materials]
class-lesson: ['spectral-data-fire-r']
permalink: /course-materials/earth-analytics/week-6/calculate-dNBR-R-Landsat/
nav-title: 'dNBR with Landsat'
week: 6
sidebar:
  nav:
author_profile: false
comments: true
order: 6
---


{% include toc title="In This Lesson" icon="file-text" %}

<div class='notice--success' markdown="1">

## <i class="fa fa-graduation-cap" aria-hidden="true"></i> Learning Objectives

After completing this tutorial, you will be able to:

* Calculate `dNBR` in `R`
* Be able to describe how the `dNBR` index is used to quantify fire severity.

## <i class="fa fa-check-square-o fa-2" aria-hidden="true"></i> What you need

You will need a computer with internet access to complete this lesson and the
data for week 6 of the course.

{% include/data_subsets/course_earth_analytics/_data-week6-7.md %}
</div>

As mentioned in the previous lesson, we can use NBR to map the extent and severity of a fire.
Let's explore creating NBR using Landsat data.

## Calculate dNBR using Landsat data

First, let's setup our spatial packages.

```{r load-packages, warning=FALSE, message=FALSE, results='hide'}
# load spatial packages
library(raster)
library(rgdal)
library(rgeos)
library(RColorBrewer)
# turn off factors
options(stringsAsFactors = F)
```

Next, we open up our landsat data and create a spatial raster stack.


```{r create-landsat-stack }
# create stack
all_landsat_bands_pre <- list.files("data/week6/Landsat/LC80340322016189-SC20170128091153/crop",
           pattern=glob2rx("*band*.tif$"),
           full.names = T) # use the dollar sign at the end to get all files that END WITH
all_landsat_bands_pre

# stack the data
landsat_stack_pre <- stack(all_landsat_bands_pre)
```

Next we calculate dNBR using the following steps:

1. Open up pre-fire data and calculate *NBR*
2. Open up the post-fire data and calculate *NBR*
3. Calculate **dNBR** (difference NBR) by subtracting post-fire NBR from pre-fire NBR.
4. Classify the dNBR raster using the classification table below.

. Note the code to do this is hidden. You will need to figure
out what bands are required to calculate NBR using Landsat.

```{r calculate-nbr, echo=FALSE, fig.cap="landsat derived NDVI plot", fig.width=7, fig.height=5}
# Landsat 8 requires bands 7 and 5
landsat_nbr_pre <- ((landsat_stack_pre[[5]] - landsat_stack_pre[[7]]) / (landsat_stack_pre[[5]] + landsat_stack_pre[[7]]))

plot(landsat_nbr_pre,
     main="Landsat derived Normalized Burn Index (NBR)\n Pre-fire - you will need to figure out the date using the Julian Day",
     axes=F,
     box=F)
```


You can export the NBR raster if you want using `writeRaster()`.

```{r export-rasters, eval=FALSE }
writeRaster(x = landsat_nbr_pre,
              filename="data/week6/outputs/landsat_nbr",
              format = "GTiff", # save as a tif
              datatype='INT2S', # save as a INTEGER rather than a float
              overwrite = T)
```


Next, we can open the post-fire landsat data to calculate post-fire NBR.

```{r create-landsat-stack-post }
all_landsat_bands_post <- list.files("data/week6/Landsat/LC80340322016205-SC20170127160728/crop",
           pattern=glob2rx("*band*.tif$"),
           full.names = T) # use the dollar sign at the end to get all files that END WITH
all_landsat_bands_post

# stack the data
landsat_stack_post <- stack(all_landsat_bands_post)
```

Then we calculate NBR on the post data - note the code here is purposefully hidden.
You need to figure out what bands to use to perform the math!

```{r calculate-nbr-post, echo=FALSE, fig.cap="landsat derived NBR post fire", fig.width=7, fig.height=5}
# Landsat 8 requires bands 7 and 5
landsat_nbr_post <- ((landsat_stack_post[[5]] - landsat_stack_post[[7]]) / (landsat_stack_post[[5]] + landsat_stack_post[[7]]))

plot(landsat_nbr_post,
     main="Landsat derived Normalized Burn Index (NBR)\n Post Fire",
     axes=F,
     box=F)
```

Finally, calculate the DIFFERENCE between the pre and post NBR!!

```{r, fig.cap="Difference NBR map", fig.width=7, fig.height=5}
# calculate difference
landsat_nbr_diff <- landsat_nbr_pre - landsat_nbr_post
plot(landsat_nbr_diff,
     main="Difference NBR map \n Pre minus post Cold Springs fire",
     axes=F, box=F)
```

When you have calculated dNBR or the difference in NBR pre minus post fire,
classify the output raster using the `classify()` function and the classes below.

| SEVERITY LEVEL  | | dNBR RANGE |
|------------------------------|
| Enhanced Regrowth | | > -.1 |
| Unburned       |  | -.1 to +.1 |
| Low Severity     | | +.1 to +.27 |
| Moderate Severity  | | +.27 to +.66 |
| High Severity     |  |  > +1.3 |

NOTE: your min an max values for NBR may be slightly different from the table
shown above! If you have a smaller min value (< -700) then adjust your first class
to that smallest number. If you have a largest max value (>1300) then adjust
your last class to that largest value in your data.

Alternatively, you can use the `Inf` to specify the smallest `-Inf` and largest
`Inf` values.


```{r classify-output, echo=F}
# create classification matrix
reclass <- c(-Inf, -.1, 1,
             -.1, .1, 2,
             .1, .27, 3,
             .27, .66, 4,
             .66, Inf , 5)
# reshape the object into a matrix with columns and rows
reclass_m <- matrix(reclass,
                ncol=3,
                byrow=TRUE)

nbr_classified <- reclassify(landsat_nbr_diff,
                     reclass_m)
the_colors = c("seagreen4", "seagreen1", "ivory1", "palevioletred1", "palevioletred4")
```


Your classified map should look something like:

```{r classify-output-plot, echo=FALSE, fig.cap="classified NBR output", fig.width=7, fig.height=5}
# mar bottom, left, top and right
par(xpd = F, mar=c(0,0,2,5))
plot(nbr_classified,
     col=the_colors,
     legend=F,
     axes=F,
     box=F,
     main="Landsat NBR - Cold Spring fire site \n Add date of the data here")
par(xpd = TRUE)
legend(nbr_classified@extent@xmax-100, nbr_classified@extent@ymax,
       c("Enhanced Regrowth", "Unburned", "Low Severity", "Moderate Severity", "High Severity"),
       fill=the_colors,
       cex=.9,
       bty="n")

```

## Compare to fire boundary

As an example to see how our fire boundary relates to the boundary that we've
identified using MODIS data, we can create a map with both layers. I'm using
the shapefile in the folder:

`data/week6/vector_layers/fire-boundary-geomac/co_cold_springs_20160711_2200_dd83.shp`

Add fire boundary to map.

```{r classify-output-plot2, echo=FALSE, fig.cap="classified NBR output", results='hide', fig.width=7, fig.height=5}
fire_boundary <- readOGR("data/week6/vector_layers/fire-boundary-geomac/co_cold_springs_20160711_2200_dd83.shp")
# reproject shapefile
fire_boundary_utm <- spTransform(fire_boundary, crs(nbr_classified))


# mar bottom, left, top and right
par(mar=c(0,0,2,5))
plot(nbr_classified,
     col=the_colors,
     legend=F,
     axes=F,
     box=F,
     main="Landsat NBR - Cold Spring fire site \n Add date of the data here")
plot(fire_boundary_utm, add=T,
     lwd=5)
legend(nbr_classified@extent@xmax-100, nbr_classified@extent@ymax,
       c("Enhanced Regrowth", "Unburned", "Low Severity", "Moderate Severity", "High Severity", "Fire boundary"),
       col= "black",
       pt.bg=rev(the_colors),
       pch=c(22, 22, 22, 22, 22, NA),
       lty = c(NA, NA, NA, NA, NA, 1),
       cex=.8,
       bty="n",
       pt.cex=c(1.75),
       xpd = TRUE)

```


```{r dev-off1, echo=FALSE, warning=FALSE, message=FALSE, results="hide"}
dev.off()
```

Make it look a bit nicer using a colorbrewer palette. I used the
`RdYlGn` palette:

`brewer.pal(5, 'RdYlGn')`


```{r classify-output-plot3, echo=FALSE, fig.cap="classified NBR output", fig.width=7, fig.height=5}
# look at colors
# display.brewer.all()
the_colors <- rev(brewer.pal(5, 'RdYlGn'))
# mar bottom, left, top and right
par(mar=c(0,0,2,5))
plot(nbr_classified,
     col=the_colors,
     legend=F,
     axes=F,
     box=F,
     main="Landsat NBR - Cold Spring fire site \n Add date of the data here")
     plot(fire_boundary_utm, add=T,
          lwd=5)
legend(nbr_classified@extent@xmax-50, nbr_classified@extent@ymax,
       c("Enhanced Regrowth", "Unburned", "Low Severity", "Moderate Severity", "High Severity", "Fire boundary"),
       col= "black",
       pt.bg = the_colors,
       pch=c(22, 22, 22, 22, 22, NA),
       lty = c(NA, NA, NA, NA, NA, 1),
       cex=.8,
       bty="n",
       pt.cex=c(1.75),
       xpd=T)
```

Note that you will have to figure out what date these data are for! I purposefully
didn't include it in the title of this map.

```{r dev-off, echo=FALSE, warning=FALSE, message=FALSE, results="hide", fig.width=7, fig.height=5}
dev.off()
```


```{r view-bar, warning=FALSE, fig.cap="plot barplot of fire severity values"}
barplot(nbr_classified,
        main="Distribution of Classified NBR Values",
        col=the_colors)
```

Add labels to your barplot!

```{r view-barplot1, warning=FALSE, fig.cap="plot barplot of fire severity values with labels"}
barplot(nbr_classified,
        main="Distribution of Classified NBR Values",
        col=the_colors,
        names.arg = c("Enhanced \nRegrowth", "Unburned", "Low \n Severity", "Moderate \n Severity", "High \nSeverity"))
```


<div class="notice--warning" markdown="1">

## <i class="fa fa-pencil-square-o" aria-hidden="true"></i> Optional challenge - NBR using MODIS

The table below shows the band ranges for the MODIS sensor. We know that the
NBR index will work with any multispectral sensor with a NIR
band between 760 - 900 nm and a SWIR band between 2080 - 2350 nm.
What bands should we use to calculate NBR using MODIS?

| Band | Wavelength range (nm) | Spatial Resolution (m) | Spectral Width (nm)|
|-------------------------------------|------------------|--------------------|----------------|
| Band 1 - red | 620 - 670 | 250 | 2.0 |
| Band 2 - near infrared | 841 - 876 | 250 | 6.0 |
| Band 3 -  blue/green | 459 - 479 | 500 | 6.0 |
| Band 4 - green | 545 - 565 | 500 | 3.0 |
| Band 5 - near infrared  | 1230 – 1250 | 500 | 8.0  |
| Band 6 - mid-infrared | 1628 – 1652 | 500 | 18 |
| Band 7 - mid-infrared | 2105 - 2155 | 500 | 18 |

</div>


<div class="notice--info" markdown="1">

## Additional Resources

* <a href="http://gsp.humboldt.edu/olm_2015/Courses/GSP_216_Online/lesson5-1/NBR.html" target="_blank">Humboldt GSP Course online NBR lesson</a>

</div>
