---
layout: single
title: "dNBR with MODIS in R"
excerpt: ". "
authors: ['Megan Cattau', 'Leah Wasser']
modified: '`r format(Sys.time(), "%Y-%m-%d")`'
category: [course-materials]
class-lesson: ['spectral-data-fire-r']
permalink: /course-materials/earth-analytics/week-6/calculate-dNBR-R-Landsat-MODIS/
nav-title: 'dNBR with Landsat & MODIS'
week: 6
sidebar:
  nav:
author_profile: false
comments: true
order: 6
---


{% include toc title="In This Lesson" icon="file-text" %}

<div class='notice--success' markdown="1">

## <i class="fa fa-graduation-cap" aria-hidden="true"></i> Learning Objectives

After completing this tutorial, you will be able to:

* Calculate `dNBR` in `R`
* Be able to describe how the `dNBR` index is used to quantify fire severity.

## <i class="fa fa-check-square-o fa-2" aria-hidden="true"></i> What you need

You will need a computer with internet access to complete this lesson and the
data for week 6 of the course.

[<i class="fa fa-download" aria-hidden="true"></i> Download Week 6 Data (~500 MB)](<a href="https://ndownloader.figshare.com/files/7677208){:data-proofer-ignore='' .btn }
</div>

As mentioned previously, we can use NBR to map the extent and severity of a fire.
Let's explore using Landsat & MODIS data to calculate NBR. In the example below, we are using
the MODIS product **mod09GA** which we downloaded from Earth Explorer. This product
contains 7 bands including ones in the SWIR and NIR region of the spectrum which we
require to calculate NBR.

## Calculate dNBR using Landsat data

Let's get started with calculating dNBR using MODIS data.
First, let's setup our spatial packages.

```{r load-packages, warning=F, message=F, results='hide'}
# load spatial packages
library(raster)
library(rgdal)
library(rgeos)
library(RColorBrewer)
# turn off factors
options(stringsAsFactors = F)
```

First, we open up our landsat data and create a spatial raster stack.


```{r create-landsat-stack }
all_landsat_bands <- list.files("data/week6/Landsat/LC80340322016205-SC20170127160728/crop",
           pattern=glob2rx("*band*.tif$"),
           full.names = T) # use the dollar sign at the end to get all files that END WITH
all_landsat_bands

# stack the data
landsat_stack_csf <- stack(all_landsat_bands)
```

Next we calculate NBR using the following steps:

1. Open up pre-fire data and calculate NBR
2. Open up the post-fire data and calculate NBR
3. Calculate dNBR (difference NBR) by subtracting post-fire NBR from pre-fire NBR.
4. 

. Note the code to do this is hidden. You will need to figure
out what bands are required to calculate NBR using Landsat.

```{r calculate-nbr, echo=F, fig.cap="landsat derived NDVI plot"}
# Landsat 8 requires bands 7 and 5
landsat_nbr <- ((landsat_stack_csf[[5]] - landsat_stack_csf[[7]]) / (landsat_stack_csf[[5]] + landsat_stack_csf[[7]])) * 1000

plot(landsat_nbr,
     main="Landsat derived NBR\n 23 July 2016",
     axes=F,
     box=F)
```

When you have calculated NBR - classify the output raster using the `classify()`
function and the classes below.

| SEVERITY LEVEL  | | dNBR RANGE |
|------------------------------|
| Enhanced Regrowth | | -700 to  -100 |
| Unburned       |  | -100 to +100 |
| Low Severity     | | +100 to +270 |
| Moderate Severity  | | +270 to +660 |
| High Severity     |  | +660 to +1300 |

NOTE: your min an max values for NBR may be slightly different from the table
shown above! If you have a smaller min value (< -700) then adjust your first class
to that smallest number. If you have a largest max value (>1300) then adjust
your last class to that largest value in your data.

Alternatively, you can set those values to NA if you think they are outside of
the valid range of NBR (in this case they are not).

```{r classify-output, echo=F}
# create classification matrix
reclass <- c(-700, -100, 1,
             -100, 100, 2,
             100, 270, 3,
             270, 660, 4,
             660, 1300, 5)
# reshape the object into a matrix with columns and rows
reclass_m <- matrix(reclass,
                ncol=3,
                byrow=TRUE)

nbr_classified <- reclassify(landsat_nbr,
                     reclass_m)
the_colors = c("palevioletred4","palevioletred1","ivory1","seagreen1","seagreen4")
```

You can export the rasters if you want.

```{r export-rasters, eval=F}
writeRaster(x = nbr_classified,
              filename="data/week6/outputs/nbr_classified.tif",
              format = "GTiff", # save as a tif
              datatype='INT2S', # save as a INTEGER rather than a float
              overwrite = T)

writeRaster(x = landsat_nbr,
              filename="data/week6/outputs/landsat_nbr",
              format = "GTiff", # save as a tif
              datatype='INT2S', # save as a INTEGER rather than a float
              overwrite = T)
```

Your classified map should look something like:

```{r classify-output-plot, echo=F, fig.cap="classified NBR output"}
# mar bottom, left, top and right
par(xpd = F, mar=c(0,0,2,5))
plot(nbr_classified,
     col=the_colors,
     legend=F,
     axes=F,
     box=F,
     main="Landsat NBR - Cold Spring fire site \n Add date of the data here")
par(xpd = TRUE)
legend(nbr_classified@extent@xmax-100, nbr_classified@extent@ymax,
       c("Enhanced Regrowth", "Unburned", "Low Severity", "Moderate Severity", "High Severity"),
       fill=rev(the_colors),
       cex=.9,
       bty="n")

```

## Compare to fire boundary

As an example to see how our fire boundary relates to the boundary that we've
identified using MODIS data, we can create a map with both layers. I'm using
the shapefile in the folder:

`data/week6/vector_layers/fire-boundary-geomac/co_cold_springs_20160711_2200_dd83.shp`

Add fire boundary to map.

```{r classify-output-plot2, echo=F, fig.cap="classified NBR output", results='hide'}
fire_boundary <- readOGR("data/week6/vector_layers/fire-boundary-geomac/co_cold_springs_20160711_2200_dd83.shp")
# reproject shapefile
fire_boundary_utm <- spTransform(fire_boundary, crs(nbr_classified))


# mar bottom, left, top and right
par(xpd = F, mar=c(0,0,2,5))
plot(nbr_classified,
     col=the_colors,
     legend=F,
     axes=F,
     box=F,
     main="Landsat NBR - Cold Spring fire site \n Add date of the data here")
plot(fire_boundary_utm, add=T,
     lwd=5)
par(xpd = TRUE)
legend(nbr_classified@extent@xmax-100, nbr_classified@extent@ymax,
       c("Enhanced Regrowth", "Unburned", "Low Severity", "Moderate Severity", "High Severity", "Fire boundary"),
       col=c(rev(the_colors), "black"),
       pch=c(15,15, 15, 15, 15,NA),
       lty = c(NA, NA, NA, NA, NA, 1),
       cex=.8,
       bty="n",
       pt.cex=c(1.75))
legend(nbr_classified@extent@xmax-100, nbr_classified@extent@ymax,
       c("Enhanced Regrowth", "Unburned", "Low Severity", "Moderate Severity", "High Severity", "Fire boundary"),
       col=c("black"),
       pch=c(22, 22, 22, 22, 22, NA),
       lty = c(NA, NA, NA, NA, NA, 1),
       cex=.8,
       bty="n",
       pt.cex=c(1.75))

```


```{r dev-off1, echo=F, warning=F, message=F, results="hide"}
dev.off()
```

Make it look a bit nicer using a colobrewer palette. I used the
`RdYlGn` palette:

`brewer.pal(5, 'RdYlGn')`

I also did a bit of legend trickery to get a box with a fill. There's probably
a better way to do this!

```{r new-legend, eval=F}

legend(nbr_classified@extent@xmax-100, nbr_classified@extent@ymax,
       c("Enhanced Regrowth", "Unburned", "Low Severity", "Moderate Severity", "High Severity", "Fire boundary"),
       col=c(rev(the_colors), "black"),
       pch=c(15,15, 15, 15, 15,NA),
       lty = c(NA, NA, NA, NA, NA, 1),
       cex=.8,
       bty="n",
       pt.cex=c(1.75))
legend(nbr_classified@extent@xmax-100, nbr_classified@extent@ymax,
       c("Enhanced Regrowth", "Unburned", "Low Severity", "Moderate Severity", "High Severity", "Fire boundary"),
       col=c("black"),
       pch=c(22, 22, 22, 22, 22, NA),
       lty = c(NA, NA, NA, NA, NA, 1),
       cex=.8,
       bty="n",
       pt.cex=c(1.75))
```

```{r classify-output-plot3, echo=F, fig.cap="classified NBR output"}
# look at colors
# display.brewer.all()
the_colors <- brewer.pal(5, 'RdYlGn')
# mar bottom, left, top and right
par(xpd = F, mar=c(0,0,2,5))
plot(nbr_classified,
     col=the_colors,
     legend=F,
     axes=F,
     box=F,
     main="Landsat NBR - Cold Spring fire site \n Add date of the data here")
     plot(fire_boundary_utm, add=T,
          lwd=5)
par(xpd = TRUE)
legend(nbr_classified@extent@xmax-100, nbr_classified@extent@ymax,
       c("Enhanced Regrowth", "Unburned", "Low Severity", "Moderate Severity", "High Severity", "Fire boundary"),
       col=c(rev(the_colors), "black"),
       pch=c(15,15, 15, 15, 15,NA),
       lty = c(NA, NA, NA, NA, NA, 1),
       cex=.8,
       bty="n",
       pt.cex=c(1.75))
legend(nbr_classified@extent@xmax-100, nbr_classified@extent@ymax,
       c("Enhanced Regrowth", "Unburned", "Low Severity", "Moderate Severity", "High Severity", "Fire boundary"),
       col=c("black"),
       pch=c(22, 22, 22, 22, 22, NA),
       lty = c(NA, NA, NA, NA, NA, 1),
       cex=.8,
       bty="n",
       pt.cex=c(1.75))

```

Note that you will have to figure out what date these data are for! I purposefully
didn't include it in the title of this map.

```{r dev-off, echo=F, warning=F, message=F, results="hide"}
dev.off()
```

```{r view-hist, warning=F, echo=F, fig.cap="plot hist"}
barplot(nbr_classified,
        main="Distribution of Classified NBR Values",
        col=the_colors)
```


### NBR & MODIS

Similarly the table below shows the band ranges for the MODIS sensor. What bands
should we use to calculate NBR using MODIS?

| Band | Wavelength range (nm) | Spatial Resolution (m) | Spectral Width (nm)|
|-------------------------------------|------------------|--------------------|----------------|
| Band 1 - red | 620 - 670 | 250 | 2.0 |
| Band 2 - near infrared | 841 - 876 | 250 | 6.0 |
| Band 3 -  blue/green | 459 - 479 | 500 | 6.0 |
| Band 4 - green | 545 - 565 | 500 | 3.0 |
| Band 5 - near infrared  | 1230 – 1250 | 500 | 8.0  |
| Band 6 - mid-infrared | 1628 – 1652 | 500 | 18 |
| Band 7 - mid-infrared | 2105 - 2155 | 500 | 18 |

<div class="notice--info" markdown="1">

## Additional Resources

* http://gsp.humboldt.edu/olm_2015/Courses/GSP_216_Online/lesson5-1/NBR.html

</div>
