---
layout: single
title: "Plot grid of plots in R. "
excerpt: " "
authors: ['Leah Wasser']
modified: '`r format(Sys.time(), "%Y-%m-%d")`'
category: [course-materials]
class-lesson: ['how-to-hints-week7']
permalink: /course-materials/earth-analytics/week-7/grid-of-plots-report/
nav-title: 'Grid of plots'
week: 7
sidebar:
  nav:
author_profile: false
comments: true
order: 3
lang-lib:
  r: []
---

{% include toc title="In This Lesson" icon="file-text" %}

<div class='notice--success' markdown="1">

## <i class="fa fa-graduation-cap" aria-hidden="true"></i> Learning Objectives

After completing this tutorial, you will be able to:

* Plot several plots using baseplot functions in a "grid" as one graphic in `R`

## <i class="fa fa-check-square-o fa-2" aria-hidden="true"></i> What you need

You will need a computer with internet access to complete this lesson and the
data for week 6/7 of the course.

{% include/data_subsets/course_earth_analytics/_data-week6-7.md %}
</div>


```{r crop-naip-imagery, message=F, warning=F}
# load libraries
library(raster)
library(rgeos)
library(rgdal)
```


```{r create-variable}
# import landsat data
all_landsat_bands <- list.files("data/week6/Landsat/LC80340322016189-SC20170128091153/crop",
           pattern=glob2rx("*band*.tif$"),
           full.names = T) # use the dollar sign at the end to get all files that END WITH

all_landsat_bands_st <- stack(all_landsat_bands)
```

### Creating a grid of plots

You can plot several plots together in the same window using baseplot. To do
this, we use the parameter value `mfrow=c(x,y)` where x is the number of rows
that you wish to have in your plot and y is the number of columns. When you plot,
R will place each plot, in order by row within the grid that you define using
`mfrow`.

Below, we have created a 2 by 2 grid of plots using `mfrow=c(2,2)` within
the `par()` function. In this example we have 2 rows and 2 columns.


```{r plot-rgb3, fig.cap="Create 2 x 2 grid of plots."}
# adjust the parameters so the axes colors are white. Also turn off tick marks.
par(mfrow=c(2,2), col.axis="white", col.lab="white", tck=0)
# plot 1
plotRGB(all_landsat_bands_st,
        r=4,g=3,b=2,
        stretch="hist",
        main="Plot 1 - RGB",
        axes=T)
box(col="white") # turn all of the lines to white

# plot 2
plotRGB(all_landsat_bands_st,
        r=5,g=4,b=3,
        stretch="hist",
        main="Plot 2 - Color Infrared (CIR)",
        axes=T)
box(col="white") # turn all of the lines to white

# plot 3
plotRGB(all_landsat_bands_st,
        r=7,g=5,b=4,
        stretch="hist",
        main="Plot 3 - Shortwave infrared",
        axes=T)
box(col="white") # turn all of the lines to white

# plot 4
plotRGB(all_landsat_bands_st,
        r=5,g=6,b=4,
        stretch="hist",
        main="Plot 4 - Land / Water",
        axes=T)
# set bounding box to white as well
box(col="white") # turn all of the lines to white

# add overall title to your layout
title("My Title", outer=TRUE)
```

Above, we added an overall title to our grid of plots using the `title()` function.
However the title is chopped of because there is not enough of a margin at the
top for it. We can adjust for this too using the `oma=` parameter argument. `oma`
sets the outside (o) margin (ma).

`oma=` argument in our `par()` function. Let's try it.


```{r plot-rgb4, fig.cap="Remove axes labels."}
# adjust the parameters so the axes colors are white. Also turn off tick marks.
par(mfrow=c(2,2), oma=c(0,0,2,0), col.axis="white", col.lab="white", tck=0)
# plot 1
plotRGB(all_landsat_bands_st,
        r=4,g=3,b=2,
        stretch="hist",
        main="Plot 1 - RGB",
        axes=T)
box(col="white") # turn all of the lines to white

# plot 2
plotRGB(all_landsat_bands_st,
        r=5,g=4,b=3,
        stretch="hist",
        main="Plot 2 - Color Infrared (CIR)",
        axes=T)
box(col="white") # turn all of the lines to white

# plot 3
plotRGB(all_landsat_bands_st,
        r=7,g=5,b=4,
        stretch="hist",
        main="Plot 3 - Shortwave infrared",
        axes=T)
box(col="white") # turn all of the lines to white

# plot 4
plotRGB(all_landsat_bands_st,
        r=5,g=6,b=4,
        stretch="hist",
        main="Plot 4 - Land / Water",
        axes=T)
# set bounding box to white as well
box(col="white") # turn all of the lines to white

# add overall title to your layout
title("My Title", outer=TRUE)
```

When you are done with plotting in a grid space, be sure to reset your plot space
using `dev.off()`.


```{r dev-off-pls2, message=F, warning=F, results="hide"}
dev.off()
```

Your homework this week should look something like this:

```{r import-boundary, echo=F, results='hide'}
# import fire overlay boundary
fire_boundary <- readOGR("data/week6/vector_layers/fire-boundary-geomac/co_cold_springs_20160711_2200_dd83.shp")
```


```{r plot-grid-naip-modis-landsat, echo=F, fig.width=7, fig.height=9, fig.cap="grid of plots" }
# NAIP
# Use stack function to read in all bands
naip_stack_csf <-
  stack("data/week6/naip/m_3910505_nw_13_1_20130926/crop/m_3910505_nw_13_1_20130926_crop.tif")

# transform the boundary
fire_boundary_utm <- spTransform(fire_boundary,
                                 CRS=crs(naip_stack_csf))

## export data
# writeOGR(fire_boundary_utm,
#           dsn="data/week6/vector_layers/fire-boundary-geomac",
#           layer="co_cold_springs_20160711_2200_utm13n",
#           driver="ESRI Shapefile",
#           overwrite_layer = T)

par(mfrow=c(3,1))

# adjust the parameters so the axes colors are white. Also turn off tick marks.
par(col.axis="white", col.lab="white", tck=0)
plotRGB(naip_stack_csf, 4,3,2,
        main="NAIP CIR image \n Cold Springs Site",
        ext=extent(fire_boundary_utm),
        axes=T)
box(col="white") # turn all of the lines to white
plot(fire_boundary_utm, add=T)

all_landsat_bands <- list.files("data/week6/Landsat/LC80340322016205-SC20170127160728/crop",
           pattern=glob2rx("*band*.tif$"),
           full.names = T) # use the dollar sign at the end to get all files that END WITH
all_landsat_bands_st <- stack(all_landsat_bands)

par(col.axis="white", col.lab="white", tck=0)
plotRGB(all_landsat_bands_st, 5,4,3,
        stretch="hist",
        main="landsat CIR image",
        ext=extent(fire_boundary_utm),
        axes=T)
box(col="white")
# add fire boundary
plot(fire_boundary_utm, add=T)


# modis cir imagery
# open modis bands
all_modis_bands <-list.files("data/week6/modis/reflectance/17_july_2016/crop",
           pattern=glob2rx("*sur_refl*.tif$"),
           full.names = T)

all_modis_bands_st <- stack(all_modis_bands)
# transform the boundary
fire_boundary_sin <- spTransform(fire_boundary,
                                 CRS=crs(all_modis_bands_st))

## export data
# writeOGR(fire_boundary_sin,
#           dsn="data/week6/vector_layers/fire-boundary-geomac",
#           layer="co_cold_springs_20160711_2200_sin",
#           driver="ESRI Shapefile",
#           overwrite_layer = T)

## 3 = blue, 4 = green, 1= red 2= nir
par(col.axis="white", col.lab="white", tck=0)
plotRGB(all_modis_bands_st,
        r=2, g =4, b=3,
        stretch="lin",
        main="MODIS CIR imagery",
        ext=extent(fire_boundary_sin),
        axes=T)
box(col="white")
plot(fire_boundary_sin, add=T)
```
