---
title: "Untitled"
output: html_document
---

```{r setup}
# import gdal
library(gdalUtils)
library(MODIS)
library(raster)
# set wd
setwd("~/Documents/earth-analytics")
```

# Helper functions

```{r pressure, echo=FALSE}
# load functions from source
# required MODIS and gdalUtils
source("~/Documents/github/earthlab.github.io/processing-code/landsat-modis-data/process-modis_functions.R")

################# 
# run function to crop all modis bands

# run function on a h4 file
the_file <- "data/week6/modis/reflectance/17_july_2016/MOD09GA.A2016199.h09v05.006.2016201065406.hdf"
process_modis_bands(the_file)


the_file <- "data/week6/modis/reflectance/07_july_2016/MOD09GA.A2016189.h09v05.006.2016191073856.hdf"
process_modis_bands(the_file)


# band_path <- the_file
# export_modis_h4_tifs <- function(band_path){
#   # this function exports all of the datasets in a MODIS h4 file
#   
#   # first get a list of layers
#   layer_list <- get_subdatasets(band_path)
#   # then 
# 
#   # export the first layer to tif
#   gdal_translate(band_path[2], dst_dataset = "/state_1km_1.tif")
#   state_rast <- raster("/state_1km_1.tif")
# 
# }
```

## How it works.

1. assign the h4 file name that you want to use
2. view all layers in the file

```{r how-it-works}
### file w 7 bands - view all layers in the h4 file
the_file <- "data/week6/modis/reflectance/07_july_2016/MOD09GA.A2016189.h09v05.006.2016191073856.hdf"
gdalinfo(the_file)

# Tells me what subdatasets are within my hdf4 MODIS files and makes them into a list
all_layers <- get_subdatasets(the_file)
all_layers

# export the first layer to tif
# gdal_translate(sds[1], dst_dataset = "/NPP2000.tif")

```


```{r }
# Load and plot the new .tif

rast <- raster("/NPP2000.tif")
plot(rast)


# then add the cloud bitmask
# gdal_translate(modis_july7[2], dst_dataset = "/state_1km_1.tif")



# export the state layer to tif
# this contains the cloud mask and is the second layer in the hdf file
# gdal_translate(modis_july7[2], dst_dataset = "/state_1km_1.tif")
state_rast <- raster("data/week6/modis/reflectance/07_july_2016/crop/MOD09GA.A2016189.h09v05.006.2016191073856_state_1km_1.tif")

# view state layer
hist(state_rast)
plot(state_rast)


#band1 - red
#band 3 - green
#band 4 - green

```

The cloud mask needs to be extracted from a bitmask layer. You can use the MODIS
package to extract and unpack the bitmask. 


http://modis-sr.ltdri.org/guide/MOD09_UserGuide_v1_3.pdf

Use detectBitInfo() notice the cloud shadow is at bit shift / mask 0 , 3
the shadow is at 2, 3

Binary Tranlation
00 = 0
01 = 1
10 = 2
11 = 3

IN this case, 1 and 2 are clouds
see the link above - table 16

| State | Translated Value | Cloud Condition| 
|----|
| 00 | 0 | clear |
| 01 | 1 | cloudy |
| 10 | 2 | mixed |
| 11 | 3 | not set, assumed clear |



```{r extract-masked-layers }
# we can use detect bit info to view the bit structure
detectBitInfo(state_rast)

## unpack bits  (required MODIS)

cloud_mask <- extractBits(state_rast, bitShift=0, bitMask=3,
                          decodeOnly=TRUE)


# open modis bands
all_modis_bands <-list.files("data/week6/modis/reflectance/07_july_2016/crop", 
           pattern=glob2rx("*sur_refl*.tif$"),
           full.names = T)

all_modis_bands_st <- stack(all_modis_bands)

# this is a 1000 m layer but i want a 500m layer 
cloud_mask_500m <- resample(cloud_mask,
         all_modis_bands_st)

# export mask 
writeRaster(cloud_mask_500m, 
            filename= "data/week6/modis/reflectance/07_july_2016/crop/cloud_mask_500m.tif",
              format = "GTiff", # save as a tif
              datatype='INT2S', # save as a INTEGER rather than a float
              overwrite = T)

# makeWeights(state_rast, bitShift=0, bitMask=3,
#            decodeOnly=TRUE)

# get inFo about bit layer for this particular modis product
detectBitInfo("MOD09GA")

plot(cloud_mask,
     main="Cloud mask layer")
barplot(cloud_mask,
     main="distribution of masked pixels",
     xlab="Cloud Status",
     names.arg=c("Clear", "Cloudy", "Mixed"))
```


```{r mask-stack}

```

