---
layout: single
title: "An example of creating modular code in R - Efficient scientific programming"
excerpt: "This lesson provides an example of modularizing code in R. "
authors: ['Max Joseph', 'Software Carpentry', 'Leah Wasser']
modified: '`r format(Sys.time(), "%Y-%m-%d")`'
category: [course-materials]
class-lesson: ['automating-your-science-r']
permalink: /course-materials/earth-analytics/week-8/function-example-modular-code-r/
nav-title: 'Function example'
week: 8
sidebar:
  nav:
author_profile: false
comments: true
order: 2
---


{% include toc title="In This Lesson" icon="file-text" %}

<div class='notice--success' markdown="1">

## <i class="fa fa-graduation-cap" aria-hidden="true"></i> Learning Objectives

After completing this tutorial, you will be able to:

* Describe how functions can make your code easier to read / follow

## <i class="fa fa-check-square-o fa-2" aria-hidden="true"></i> What you need

You will need a computer with internet access to complete this lesson and the
data that we already downloaded for week 6 of the course.

{% include/data_subsets/course_earth_analytics/_data-week6-7.md %}
</div>


```{r}

# set working dir
setwd("~/Documents/earth-analytics")

# load spatial packages
library(raster)
library(rgdal)
# turn off factors
options(stringsAsFactors = F)

# get list of tif files
all_landsat_bands <- list.files("data/week6/Landsat/LC80340322016189-SC20170128091153/crop",
                                pattern=glob2rx("*band*.tif$"),
                                full.names = T)

# stack the data (create spatial object)
landsat_stack_csf <- stack(all_landsat_bands)

par(col.axis="white", col.lab="white", tck=0)
# plot brick
plotRGB(landsat_stack_csf,
  r=4,g=3, b=2, 
  main="RGB Landsat Stack \n pre-fire",
  axes=T,
  stretch="hist")
box(col="white") # turn all of the lines to white



```


```{r }

# we can do the same things with functions
get_stack_bands <- function(the_dir_path, the_pattern){
  # get list of tif files
  all_landsat_bands <- list.files(the_dir_path,
                                pattern=glob2rx(the_pattern),
                                full.names = T)

  # stack the data (create spatial object)
  landsat_stack_csf <- stack(all_landsat_bands)
  return(landsat_stack_csf)
  
}



```


# Example using functions

Here's we've reduced the code by a few lines using a get bands function. Then we 
can plot like we did before. 

```{r plot-rgb, fig.cap="raster stack plot - rgb"}
# code to go here
landsat_pre_fire <- get_stack_bands(the_dir_path = "data/week6/Landsat/LC80340322016189-SC20170128091153/crop",
                the_pattern = "*band*.tif$")

par(col.axis="white", col.lab="white", tck=0)
# plot brick
plotRGB(landsat_pre_fire,
  r=4,g=3, b=2, 
  main="RGB Landsat Stack \n pre-fire",
  axes=T,
  stretch="lin")
box(col="white") # turn all of the lines to white

```


Now, what if we created a function that adjusted
all of the parameters that we wanted to set to plot an RGB image? Here we 
will require the user to send the function a stack with the bands in the order
that they want to plot the data. 

```{r create-plot-function}
# notice here i set a default stretch to blank
create_rgb_plot <-function(a_raster_stack, the_plot_title, the_stretch=""){
  # a function that plots a raster stack turning on axes and setting the parameters to 
  # white to "hide" the axes for a nice looking plot with a title
  # inputs: raster stack with multiple spectral raster layers
  # outputs
  par(col.axis="white", col.lab="white", tck=0)
  # plot raster stack
  plotRGB(a_raster_stack, 
    main=the_plot_title,
    axes=T,
    stretch=the_stretch)
  box(col="white") # turn all of the lines to white
  
}

```


```{r plot-rgb-function, fig.cap="raster stack plot - rgb"}
# code to go here
landsat_pre_fire <- get_stack_bands(the_dir_path = "data/week6/Landsat/LC80340322016189-SC20170128091153/crop",
                the_pattern = "*band*.tif$")

# stack the data in the order that you want to plot
landsat_pre_fire_RGB <- stack(landsat_pre_fire[[4]], landsat_pre_fire[[3]], landsat_pre_fire[[2]])

# plot the data 
create_rgb_plot(a_raster_stack = landsat_pre_fire_RGB,
                the_plot_title = "RGB image",
                the_stretch="hist")
```

Once our plot parameters are setup, we can use the same code to plot our data 
over and over without having to set parameters each time!

```{r}


```

