---
layout: single
title: "Working with function arguments"
excerpt: " ."
authors: ['Max Joseph', 'Software Carpentry',  'Leah Wasser']
modified: '`r format(Sys.time(), "%Y-%m-%d")`'
category: [course-materials]
class-lesson: ['automating-your-science-r']
permalink: /course-materials/earth-analytics/week-8/function-arguments-r/
nav-title: 'Function arguments'
week: 8
sidebar:
  nav:
author_profile: false
comments: true
order: 6
---


{% include toc title="In This Lesson" icon="file-text" %}

<div class='notice--success' markdown="1">

## <i class="fa fa-graduation-cap" aria-hidden="true"></i> Learning Objectives

After completing this tutorial, you will be able to:

* Define the purpose of a function argument.
* Use default vs. required function arguments in a function.

## <i class="fa fa-check-square-o fa-2" aria-hidden="true"></i> What you need

You will need a computer with internet access to complete this lesson.

</div>

In the previous lessons, we have used many different functions and function
arguments to customize our code.

```{r import-raster, echo=F}
library(raster)
library(rgdal)
```

For example, we used numerous arguments to plot our data including:

1. `main` to add a title
2. `axes = FALSE` to remove the axes of our plot
3. `box = FALSE` to remove the box surrounding the plot.

In the example below, we call each argument by name and then assign it a value
based on the type of argument it is. For example the value for the `main = ` argument
is a text string which is the title that we want `R` to add to our plot.

```{r plot-ndvi, fig.cap="ndvi plot", fig.width=7, fig.height=5}
# import and plot landsat
landsat_ndvi <- raster("data/week6/outputs/landsat_ndvi.tif")
plot(landsat_ndvi,
     main = "ndvi title - rendered using a function argument",
     axes = FALSE,
     box = FALSE)
```

Function arguments allow us to customize how a function runs. For example, we can
use the `plot()` function to plot many different types of data! And we can use the
`main` argument to customize the title. We use axes and box to customize how the
plot looks. This is a powerful function as it can be used to do many different
things and is customizable in many ways that we may need / want!

## Argument order matters

Let's next talk about the order or arguments in a function. `R` has three ways
that arguments supplied by you are matched to the *formal arguments* of the
function definition:

1. **By complete name:** i.e. you type `main = ""` and `R` matches main to the argument called `main`.
2. **By order or position when you call an argument:** i.e. you call `plot(raster, "title here")`, `R` will read these two variables in the order that you provide them. This can cause the function to fail if they are not in the right order!
3. **By partial name:** (matching on initial *n* characters of the argument name) - we are not going to review this in class. Beware using this "feature".

Arguments are matched in the manner outlined above.

* R first tries to find arguments according to the complete name,
* then by partial matching of names,
* and finally by position.

With that in mind, let's look at the help for `read.csv()`:

```{r, eval=FALSE}
# view help for the csv function
?read.csv
```

There's a lot of information available in the help. The the most important part
is the first couple of lines:

```{r, eval=FALSE}
read.csv(file, header = TRUE, sep = ",", quote = "\"",
         dec = ".", fill = TRUE, comment.char = "", ...)
```

This tells us that `read.csv()` has one argument, `file`, that doesn't have a
default value, and six other arguments that do have a default value.

Now we understand why the following code returns an error:

```{r, results="hide", error = TRUE}
precip_data <- read.csv(FALSE, "data/week2/precipitation/precip-boulder-aug-oct-2013.csv")
```

The code above fails because `FALSE` is assigned to `file` and the filename is assigned to
the argument `header`.

## Default function arguments

We have passed arguments to functions in two ways:

1. Directly: `plot(landsat_ndvi)`,
2. and by name: `read.csv(file = "data/week2/precipitation/precip-boulder-aug-oct-2013.csv", header = FALSE)`.

We can pass the arguments to `read.csv` without naming them if they are in the
order that R expects.

```{r, eval=F}
precip_data <- read.csv("data/week2/precipitation/precip-boulder-aug-oct-2013.csv",
                FALSE)
```

However, the position of the arguments matter if they are not named.
Does the code below return an error?

```{r, eval=F}
# import csv
precip_data <- read.csv(header = FALSE,
                file = "data/week2/precipitation/precip-boulder-aug-oct-2013.csv")

```

But this code below doesn't work. Make sense?

```{r, error = TRUE}
dat <- read.csv(FALSE,
                "data/week2/precipitation/precip-boulder-aug-oct-2013.csv")

```
