---
layout: single
title: "Understand the function environment & function arguments in R"
excerpt: "This lesson introduces the function environment and documenting functions in R. When you run a function intermediate variables are not stored in the global environment. This not only saves memory on your computer but also keeps our environment clean, reducing the risk of conflicting variables."
authors: ['Max Joseph', 'Software Carpentry', 'Leah Wasser']
modified: '`r format(Sys.time(), "%Y-%m-%d")`'
category: [course-materials]
class-lesson: ['automating-your-science-r']
permalink: /course-materials/earth-analytics/week-8/function-documentation-environment-r/
nav-title: 'Function documentation & environment'
week: 8
sidebar:
  nav:
author_profile: false
comments: true
order: 5
---


{% include toc title="In This Lesson" icon="file-text" %}

<div class='notice--success' markdown="1">

## <i class="fa fa-graduation-cap" aria-hidden="true"></i> Learning Objectives

After completing this tutorial, you will be able to:

* Document a function in R describing the function purpose, inputs, outputs and associated structures
* Describe what happens to intermediate variables processed during a function call

## <i class="fa fa-check-square-o fa-2" aria-hidden="true"></i> What you need

You will need a computer with internet access to complete this lesson.

</div>

In the last lesson, we learned how to create a function in `R`. We talked about
how functions are efficient ways to reduce variables in your global environment.
In this lesson we will explore that further. We will also explore function arguments.

## The function environment is self-contained

There are many benefits to using functions in your code including simplicity,
and expressiveness. However, functions also save memory by keeping objects out
of your global environment.

The function environment is self-contained. What this  means is that when you
run a function, it does not create intermediate variables in your global environment.

For example, in the previous lessons, we created a function called
`fahr_to_celsius`. Within that function, we created two variables:

1. `kelvin`
2. `celsius`

Run the function below. Then call it `fahr_to_celsius(15)`. Look closely at
your global environment in `R`. Do you see the variables `temp_k` or `result`
in your list of variables in R Studio?

```{r funct-environment}
fahr_to_celsius <- function(fahr) {
  kelvin <- fahr_to_kelvin(fahr)
  celsius <- kelvin_to_celsius(kelvin)
  celsius
}

fahr_to_celsius(15)

```

When we run the function above, it creates a new temporary environment where it runs
the steps required to complete the tasks specified in the function. However,
the variables defined by each intermediate steps are not retained in your global
environment. These variables only exist within the function environment. This
is convenient for us as programmers because each variable that you create consumes
memory on your computer. It also reduces the "clutter" associated with too many
variables in your global environment which could conflict further down in your
code. For example you may have another variable called "celsius" in your code
code further down.


## Documentation

Now that we understand the function environment, let's talk about documenting
our functions. This documentation is important to

1. Remind ourselves later what the function does
1. Show how to use the function, and
1. To help anyone else looking at our code understand what **we think** the function does.

Note that our written documentation can at best describe what **we think** the function does, because ultimately the code itself is the only true documentation for the what the function **actually** does.
Beware!

A common way to add documentation in software is to add comments to your function
that specify

1. What does this function do?
2. What are the arguments (inputs) to the function, and what are these supposed to be (e.g., what class are they? Character, numeric, logical?)
3. What does the function return, and what kind of object is it?

Like this:

```{r document-function}
fahr_to_celsius <- function(fahr) {
  # convert temperature in fahrenheit to celsius
  # args: temperature in degrees F (numeric)
  # returns: temperature in degrees celsius (numeric)
  kelvin <- fahr_to_kelvin(fahr)
  celsius <- kelvin_to_celsius(kelvin)
  celsius
}
```

> ## Writing Documentation
>
> Formal documentation for R functions that you see when you access the help in
> R is written in separate `.Rd` using a
> markup language similar to LaTeX. You see the result of this documentation
> when you look at the help file for a given function, e.g. `?read.csv`.
> The `roxygen2` package allows `R` coders to write documentation alongside
> the function code and then process it into the appropriate `.Rd` files.
> You should consider switching this more formal method of writing documentation
> when you start working on more complicated `R` projects. Or if you aspire to
> write packages in R!
>
> * <a href="http://www.latex-project.org/" target="_blank">More on LaTeX</a>
> * <a href="http://cran.r-project.org/web/packages/roxygen2/vignettes/rd.html" target="_blank"> More on roxygen2</a>
{: .notice }
