---
layout: single
title: "Subset & aggregate time series precipitation data in R"
excerpt: "This lesson walks aggregating time series data in R."
authors: ['Leah Wasser', 'Mariela Perignon']
modified: '`r format(Sys.time(), "%Y-%m-%d")`'
category: [course-materials]
class-lesson: ['hw-ggplot2-r']
week: 2
permalink: /course-materials/earth-analytics/week-2/aggregate-time-series-data-r/
nav-title: 'GRAD / BONUS: aggregate data'
sidebar:
  nav:
author_profile: false
comments: true
order: 5
---

{% include toc title="In This Lesson" icon="file-text" %}


Bonus / graduate activity. In this lesson, you will PLOT precipitation data in R.
However, these data were collected over several decades and sometimes there are
multiple data points per day. The data are also not cleaned. You will find
heading names that may not be meaningful, and other issues with the data.

This lesson provides the basic skills that you need to create a plot of daily
precipitation, for 30 years surrounding the 2013 flood. You will use the skills
that you learned in the previous lessons, coupled with the skills in this lesson
to process the data.

<div class='notice--success' markdown="1">

## <i class="fa fa-graduation-cap" aria-hidden="true"></i> Learning Objectives

After completing this tutorial, you will be able to:

* Aggregate data by a day in R
* View names and rename columns in a dataframe.

### Things You'll Need To Complete This Lesson

Please be sure you have the most current version of R and, preferably,
`RStudio` to write your code.

 **R Skill Level:** Intermediate - To succeed in this tutorial, you will need to
have basic knowledge for use of the `R` software program.

### R Libraries to Install:

* **ggplot2:** `install.packages("ggplot2")`
* **plotly:** `install.packages("dplyr")`

#### Data Download

If you haven't already downloaded this data (from the previous lesson), do so now.

[<i class="fa fa-download" aria-hidden="true"></i> Download Precipitation Data](https://ndownloader.figshare.com/files/7426738){:data-proofer-ignore='' .btn }

</div>


## Work with Precipitation Data

## R Libraries

To get started, load the `ggplot2` and `dplyr` libraries, setup your working
directory and set `stringsAsFactors` to FALSE using `options()`.

```{r load-libraries, echo=F }

# setwd("working-dir-path-here")

# load packages
library(ggplot2)
library(dplyr)

options(stringsAsFactors = FALSE)

```


## Import Precipitation Data

We will use the `805333-precip-daily-1948-2013.csv` dataset for this assignment.
in this analysis. This dataset contains the precipitation values collected daily
from the COOP station 050843 in Boulder, CO for 1 January 2003 through 31 December 2013.

Import the data into R and then view the data structure.

```{r import-precip, echo=F}

#precip_file <- "data/flood-co-2013/precip/805325-precip_daily_2003-2013.csv"
precip_file <- "data/week2/precipitation/805333-precip-daily-1948-2013.csv"

# import precip data into R data.frame
precip.boulder <- read.csv(precip_file,
                           header = TRUE)

# view first 6 lines of the data
head(precip.boulder)

# view structure of data
str(precip.boulder)

```

## About the Data

The structure of the data are similar to what you saw in previous lessons. HPCP
is the total precipitation given in inches, recorded
for the hour ending at the time specified by DATE. There is a designated missing
data value of 999.99. Note that hours with no precipitation are not recorded.


The metadata for this file is located in your week2 directory:
`PRECIP_HLY_documentation.pdf` file that can be downloaded along with the data.
(Note, as of Sept. 2016, there is a mismatch in the data downloaded and the
documentation. The differences are in the units and missing data value:
inches/999.99 (standard) or millimeters/25399.75 (metric)).

### NoData Values

Next, check out the data. Are there no data values? If so, make sure to adjust your
data import code above to account for no data values. Then determine how many no
data values you have in your dataset.

``` {r no-data-values-hist, echo=F, fig.cap="histogram of data"}
#plot histogram
hist(precip.boulder$HPCP, main ="Are there NA values?")

precip.boulder <- read.csv(precip_file,
                           header = TRUE, na.strings = 999.99)

hist(precip.boulder$HPCP, main ="This looks better after the reimporting with\n no data values specified", xlab="Precip (inches)", ylab="Frequency")


```


```{r how-many-na}
print("how many NA values are there?")
sum(is.na(precip.boulder))
```

### Convert Date and Time

Compared to the previous lessons, notice that we now have date & time in our date field.
To deal with both date and time, we use the `as.POSIXct()` method rather than
as.date which we used previously. The syntax to convert to POSIXct is similar to
what we used previously, but now, we will add the hour (H) and minute (M) to the
format argument as follows:

`as.POSIXct(column-you-want-to-convert-here, format="%Y%m%d %H:%M")`

```{r convert-date, echo=F, results='hide'}

# convert to date/time and retain as a new field
precip.boulder$DATE <- as.POSIXct(precip.boulder$DATE,
                                  format="%Y%m%d %H:%M")
                                  # date in the format: YearMonthDay Hour:Minute

# double check structure
str(precip.boulder$DATE)

```

* For more information on date/time classes, see the NEON tutorial
<a href="http://neondataskills.org/R/time-series-convert-date-time-class-POSIX/" target="_blank"> *Dealing With Dates & Times in R - as.Date, POSIXct, POSIXlt*</a>.


## Plot Precipitation Data

Next, let's have a look at the data. Plot using `ggplot()`. Format the plot using
the colors, labels, etc that are most clear and look the best. Your plot does not
need to look like the one below!

``` {r plot-precip-hourly, echo=F, fig.cap="hourly precipitation"}

# plot the data using ggplot2
precPlot_hourly <- ggplot(precip.boulder, aes(DATE, HPCP)) +   # the variables of interest
      geom_point(stat="identity") +   # create a bar graph
      xlab("Date") + ylab("Precipitation (Inches)") +  # label the x & y axes
      ggtitle("Hourly Precipitation - Boulder Station\n 1948-2013")  # add a title

precPlot_hourly

```

## Differences in the data
Any ideas what might be causing the notable difference in the plotted data through time?

``` {r plot-precip-hourly-round, echo=F, fig.cap="hourly precipitation"}
# round precip
precip.boulder$HPCP_round <- round(precip.boulder$HPCP, digits = 1)

# plot the data using ggplot2
precPlot_hourly_round <- ggplot(precip.boulder, aes(DATE, HPCP_round)) +   # the variables of interest
      geom_point(stat="identity") +   # create a bar graph
      xlab("Date") + ylab("Precipitation (Inches)") +  # label the x & y axes
      ggtitle("Hourly Precipitation - Boulder Station\n 1948-2013")  # add a title

precPlot_hourly_round

```

It is difficult to interpret this plot which spans so many years at such a fine 
temporal scale. For our research project, we only need to explore 30 years of data.
Let's do the following:

1. Aggregate the precipitation totals (sum) by day.
2. Subset the data for 30 years (we learned how to do this in a previous lesson).

#### Aggregating and summarizing data

To aggregate data by a particular variable or time period, we can create a new column
in our dataset called day. We will take all of the values for each day and add them
using the `sum()` function. We can do all of this efficiently using dplyr mutate() function.

We use the `mutate()` function to add a new column called **day** to a new data.frame called **daily_sum_precip**. Note that we used `as.Date()` to just grab the dates rather than dates and times which are stored in the POSIX format.


```{r daily-summaries }
# use dplyr
daily_sum_precip <- precip.boulder %>%
  mutate(day = as.Date(DATE, format="%Y-%m-%d"))   # create a new column called day w the date


# let's look at the new column
head(daily_sum_precip$day)

```

```{r plot-daily, echo=F, warning=F, fig.cap="Daily precip plot"}
precPlot_daily1 <- ggplot(data=precip.boulder,  # the data frame
      aes(DATE, HPCP)) +   # the variables of interest
      geom_bar(stat="identity") +   # create a bar graph
      xlab("Date") + ylab("Precipitation (Inches)") +  # label the x & y axes
      ggtitle("Daily Precipitation - Boulder Station\n 2003-2013")  # add a title

precPlot_daily1

```

Next we `summarize()` the precipitation column (total_precip) - grouped by day.
What this means is that we ADD UP all of the values for each day to get a grand
total amount of precipitation each day.


``` {r daily-summ}
# use dplyr
daily_sum_precip <- precip.boulder %>%
  mutate(day = as.Date(DATE, format="%Y-%m-%d")) %>%
  group_by(day) %>% # group by the day column
  summarise(total_precip=sum(HPCP)) # calculate the SUM of all precipitation that occured on each day

# how large is the resulting data frame?
nrow(daily_sum_precip)

# view the results
head(daily_sum_precip)

# view column names
names(daily_sum_precip)

```


Now plot the daily data.

``` {r daily-prec-plot, echo=F, warning=F, fig.cap="Daily precipitation for boulder"}
# plot daily data
precPlot_daily <- ggplot(daily_sum_precip, aes(day, total_precip)) +
      geom_bar(stat="identity") +
      xlab("Date") + ylab("Precipitation (inches)") +
      ggtitle("Daily Precipitation - Boulder Station\n 2003-2013")

precPlot_daily
```


Finally, plot a temporal subsets of the data from 2000-2013. We learned how to 
do this in the previous lessons. 


Now we can easily see the dramatic rainfall event in mid-September!

<i class="fa fa-star"></i> **R Tip:** If you are using a date-time class, instead
of just a date class, you need to use `scale_x_datetime()`.
{: .notice}

#### Subset The Data

If we wanted to, we could subset this data set using the same code that we used 
previously to subset! An example of the subsetted plot is below. 

``` {r subset-data, echo=F, warning='hide', fig.cap="final precip plot daily sum"}

# use dplyr
daily_sum_precip_subset <- daily_sum_precip %>%
  filter(day >= as.Date('2012-12-31') & day <= as.Date('2013-12-31'))


# create new plot
precPlot_30yrs <- ggplot(daily_sum_precip_subset, aes(day, total_precip)) +
  geom_bar(stat="identity") +
  xlab("Date") + ylab("Precipitation (inches)") +
  ggtitle("Daily Total Precipitation 2012 - 2013 for Boulder Creek")

precPlot_30yrs

```

<div class="notice--info" markdown="1">

## Additional Resources

* <a href="http://stackoverflow.com/questions/11395927/how-to-subset-data-frame-by-weeks-and-then-sum" target="_blank">How to subset data by weeks</a>
</div>


