---
layout: single
title: "Plot with GGPLOT"
excerpt: 'This lesson walks through using GGPLOT2 to plot data.'
authors: ['Leah Wasser', 'Data Carpentry']
modified: '`r format(Sys.time(), "%Y-%m-%d")`'
category: [course-materials]
class-lesson: ['hw-ggplot2-r']
module-description: 'This tutorial covers how to plot time series data in R using ggplot2.
  It also covers converting data stored in data/time format into an R date time class.'
module-nav-title: 'Time Series Data in R'
module-title: 'Work with Sensor Network Derived Time Series Data in R'
module-type: 'homework' 
course: "Earth Analytics"
nav-title: 'GGPLOT R'
permalink: /course-materials/earth-analytics/week-2/hw-ggplot2-r/
week: 2
sidebar:
  nav:
author_profile: false
comments: true
order: 1
---

{% include toc title="In This Lesson" icon="file-text" %}

In this tutorial, we will explore more advanced plotting techniques using `ggplot2`.

<div class='notice--success' markdown="1">

## <i class="fa fa-graduation-cap" aria-hidden="true"></i> Learning Objectives
At the end of this activity, you will be able to:

* Use the `ggplot()` plot function to create custom plots.
* Add labels to x and y axes and a title to your ggplot plot.
* Customize the colors and look of a ggplot plot.

## <i class="fa fa-check-square-o fa-2" aria-hidden="true"></i> What you need

You need `R` and `RStudio` to complete this tutorial. Also you should have
an `earth-analytics` directory setup on your computer with a `/data`
directory with it.

* [How to Setup R / RStudio](/course-materials/earth-analytics/week-1/setup-r-rstudio/)
* [Setup your working directory](/course-materials/earth-analytics/week-1/setup-working-directory/)
* [Intro to the R & RStudio Interface](/course-materials/earth-analytics/week-1/intro-to-r-and-rstudio)

</div>


In our week 1 homework, we used the quick plot function of ggplot2 to plot our data.
In this tutorial, we'll explore ggplot - which offers many more advanced plotting
features.

Let's explore the code below to create a quick plot.

```{r plot-data, fig.cap="quick plot of precip data"}

# load the ggplot2 library for plotting
library(ggplot2)

# download data from figshare
# note that we already downloaded the data to our laptops previously
# but in case you don't have it - re-download it by uncommenting the code below.
#download.file(url = "https://ndownloader.figshare.com/files/7010681",
#              destfile = "data/boulder-precip.csv")

# import data
boulder_precip <- read.csv(file="data/boulder-precip.csv")

# view first few rows of the data
head(boulder_precip)

# when we download the data we create a dataframe
# view each column of the data frame using its name (or header)
boulder_precip$DATE

# view the precip column
boulder_precip$PRECIP

# q plot stands for quick plot. Let's use it to plot our data
qplot(x=boulder_precip$DATE,
      y=boulder_precip$PRECIP)

```

## Plotting with ggplot2

`ggplot2` is a plotting package that makes it simple to create complex plots
from data in a dataframe. It uses default settings, which help to create
publication quality plots with a minimal amount of settings and tweaking.

ggplot graphics are built step by step by adding new elements.

To build a ggplot we need to:

- bind the plot to a specific data frame using the `data` argument

```{r, eval=FALSE, purl=FALSE, fig.cap="ggplot binding"}
ggplot(data = boulder_precip)

```

- define aesthetics (`aes`), by selecting the variables to be plotted and the variables to define the presentation
     such as plotting size, shape color, etc.,

```{r, eval=FALSE, purl=FALSE, fig.cap="ggplot example plot precip"}
ggplot(data = boulder_precip, aes(x = DATE, y = PRECIP))
```

- add `geoms` -- graphical representation of the data in the plot (points,
     lines, bars). To add a geom to the plot use `+` operator:

```{r first-ggplot, purl=FALSE, fig.cap="ggplot boulder precip"}
ggplot(data = boulder_precip,  aes(x = DATE, y = PRECIP)) +
  geom_point()

```

The `+` in the `ggplot2` package is particularly useful because it allows you
to modify existing `ggplot` objects. This means you can easily set up plot
"templates" and conveniently explore different types of plots, so the above
plot can also be generated with code like this:

```{r, first-ggplot-with-plus, eval=FALSE, purl=FALSE, fig.cap="first ggplot"}
# Create the plot object (nothing will render on your screen)
precip_plot <-  ggplot(data = boulder_precip,  aes(x = DATE, y = PRECIP))

# Draw the plot
precip_plot + geom_point()

```

```{r, eval=FALSE, purl=TRUE, echo=FALSE, purl=FALSE, fig.cap="2nd ggplot"}
# Create
precip_plot <-  ggplot(data = boulder_precip,  aes(x = DATE, y = PRECIP))

# Draw the plot
precip_plot + geom_point()
```


We can also apply a color to our points

```{r adding-colors, purl=FALSE, fig.cap="ggplot with blue points"}
ggplot(data = boulder_precip,  aes(x = DATE, y = PRECIP)) +
    geom_point(color = "blue")

```

And adjust the transparency.

```{r add-alpha, fig.cap="ggplot with blue points and alpha"}
ggplot(data = boulder_precip,  aes(x = DATE, y = PRECIP)) +
    geom_point(alpha=.5, color = "blue")

```


Or to color each value in the plot differently:

```{r color-by-species, purl=FALSE, fig.cap="ggplot with colored points"}
ggplot(data = boulder_precip,  aes(x = DATE, y = PRECIP)) +
    geom_point(alpha = 0.9, aes(color=PRECIP))

```


We can turn our plot into a bar plot.

```{r barplot, purl=FALSE, fig.cap="ggplot with bars"}
ggplot(data = boulder_precip,  aes(x = DATE, y = PRECIP)) +
    geom_bar(stat="identity")

```

Turn the bar outlines blue

```{r bar-color, purl=FALSE, fig.cap="ggplot with blue bars"}
ggplot(data = boulder_precip,  aes(x = DATE, y = PRECIP)) +
    geom_bar(stat="identity", color="blue")

```

Change the fill to bright green.

```{r barcolor2, purl=FALSE, fig.cap="ggplot with green bars"}
ggplot(data = boulder_precip,  aes(x = DATE, y = PRECIP)) +
    geom_bar(stat="identity", color="blue", fill="green")

```


## Adding Labels

You can add labels to your plots as well. Let's add a title, and x and y labels.
```{r add-title, fig.cap="ggplot with labels" }
ggplot(data = boulder_precip,  aes(x = DATE, y = PRECIP)) +
    geom_point(alpha = 0.9, aes(color=PRECIP)) +
    ggtitle("Precipitation - Boulder, Colorado 2013")
```

x and y labels...


```{r add-title2, fig.cap="ggplot with title and labels" }
ggplot(data = boulder_precip,  aes(x = DATE, y = PRECIP)) +
    geom_point(alpha = 0.9, aes(color=PRECIP)) +
    ggtitle("Precipitation - Boulder, Colorado 2013") +
    xlab("x label here") + ylab("y label here")

```


## More on customizing your plots

There are many different tutorials out there on customizing ggplot plots. A
few are listed below.

* <a href="http://www.datacarpentry.org/R-ecology-lesson/05-visualization-ggplot2.html" target="_blank" data-proofer-ignore=''> Data carpentry ggplot2 </a>
* <a href="http://www.cookbook-r.com/Graphs/" target="_blank">R Cookbook</a>

<div class="notice--warning" markdown="1">

## <i class="fa fa-pencil-square-o" aria-hidden="true"></i> Optional challenge

Customize the plot that we created in last weeks homework as follows:

* Change the colors of the plot
* Plot the data using points `geom_point()` AND bars `geom_bar()`
* Add a title to your plot and x and y axis labels.

</div>
