---
layout: single
title: "Use MODIS & Landsat to quantify fire in R"
excerpt: " "
authors: ['Megan Cattau', 'Leah Wasser']
modified: '`r format(Sys.time(), "%Y-%m-%d")`'
category: [course-materials]
class-lesson: ['spectral-data-fire-2-r']
permalink: /course-materials/earth-analytics/week-6/intro-spectral-data-r/
nav-title: 'Intro spectral data'
module-title: 'Use MODIS & Landsat to quantify fire in R'
module-description: ''
module-nav-title: 'Fire / spectral remote sensing data - in R'
module-type: 'class'
week: 6
sidebar:
  nav:
author_profile: false
comments: true
order: 1
output: html_document
---

{% include toc title="In This Lesson" icon="file-text" %}

<div class='notice--success' markdown="1">

## <i class="fa fa-graduation-cap" aria-hidden="true"></i> Learning Objectives

After completing this tutorial, you will be able to:

* 

## <i class="fa fa-check-square-o fa-2" aria-hidden="true"></i> What you need

You will need a computer with internet access to complete this lesson and the
data for week 6 of the course.

[<i class="fa fa-download" aria-hidden="true"></i> Download Week 6 Data (~500 MB)](https://ndownloader.figshare.com/files/7636975){:data-proofer-ignore='' .btn }



</div>



## 

```{r import-libraries }
# import spatial packages
library(raster)
library(rgdal)
library(rgeos)
# turn off factors
options(stringsAsFactors = F)

```



```{r }
all_landsat_bands <- list.files("data/week6/Landsat/LC80340322016189-SC20170128091153/crop",
           pattern=glob2rx("*band*.tif$"),
           full.names = T) # use the dollar sign at the end to get all files that END WITH

all_landsat_bands_st <- stack(all_landsat_bands)
plotRGB(all_landsat_bands_st,
        r=4, g=3, b=2,
        stretch="hist")

# apply shadow mask
cloud_mask_189 <- raster("data/week6/Landsat/LC80340322016189-SC20170128091153/crop/LC80340322016189LGN00_cfmask_crop.tif")
plot(cloud_mask_189)
cloud_mask_189_conf <- raster("data/week6/Landsat/LC80340322016189-SC20170128091153/crop/LC80340322016189LGN00_cfmask_conf_crop.tif")
plot(cloud_mask_189_conf)

# Create mask 
cloud_mask_189[cloud_mask_189 > 0] <- NA
plot(cloud_mask_189)

# mask the stack
all_landsat_bands_mask <- mask(all_landsat_bands_st, mask = cloud_mask_189)
plotRGB(all_landsat_bands_mask,
        r=4, g=3, b=2,
        stretch="hist")




## Create a function to calculate a veg index
get_veg_index <- function(band1, band2){
  # this function calculates the normalize difference between two bands
  # output: a new raster with index values bewteen -1 and 1
  new_index <- (band2 - band1) / (band2 + band1)
  return(new_index)
}

# calculate NBR

landsat_nbr <- overlay(all_landsat_bands_mask[[4]], all_landsat_bands_mask[[5]], 
                       fun=get_veg_index)
plot(landsat_nbr)
```


## good opportunity to learn about projections
let them plot modis, let them calculate NBR
The overlay the fire boundary -- they will have to 

1. figure out band combinations
2. reproject


```{r work-with-modis}
# open modis bands
all_modis_bands <-list.files("data/week6/modis/reflectance/07_july_2016/crop", 
           pattern=glob2rx("*sur_refl*.tif$"),
           full.names = T)

all_modis_bands_st <- stack(all_modis_bands)
## 3 = blue, 4 = green, 1= red 2= nir
plotRGB(all_modis_bands_st, 
        r=1, g =4, b=3,
        stretch="lin")

# view fire overlay boundary
fire_boundary <- readOGR("data/week6/vector_layers/fire-boundary-geomac/co_cold_springs_20160711_2200_dd83.shp")
fire_boundary_sin <- spTransform(fire_boundary,
                                 CRS=crs(all_modis_bands_st))

# export as sinusoidal
writeOGR(fire_boundary_sin,
         dsn = "data/week6/vector_layers/fire-boundary-geomac",
         layer="co_cold_springs_20160711_2200_sin",
         driver="ESRI Shapefile", 
         overwrite_layer=TRUE)


# plot(fire_boundary_sin, lwd=100)


```

## NOTE they don't have this cloud layer in their data. 

| State | Translated Value | Cloud Condition| 
|----|
| 00 | 0 | clear |
| 01 | 1 | cloudy |
| 10 | 2 | mixed |
| 11 | 3 | not set, assumed clear |


```{r create & apply mask}
# import cloud mask
cloud_mask_7July <- raster("data/week6/modis/reflectance/07_july_2016/crop/cloud_mask_500m.tif")
cloud_mask_7July[cloud_mask_7July > 0] <- NA
plot(cloud_mask_7July)

all_modis_bands_st_mask <- mask(all_modis_bands_st,
                                cloud_mask_7July)

## 3 = blue, 4 = green, 1= red 2= nir
plotRGB(all_modis_bands_st, 
        r=1, g =4, b=3,
        stretch="lin")

## 3 = blue, 4 = green, 1= red 2= nir
plotRGB(all_modis_bands_st_mask, 
        r=1, g =4, b=3,
        stretch="lin")

fire_bound_sin <- readOGR("data/week6/vector_layers/fire-boundary-geomac/co_cold_springs_20160711_2200_sin.shp")
plot(fire_bound_sin, add=T, col="yellow", lwd=1)
```



< -0.25 
High post-fire regrowth 
-0.25 to -0.1 
Low post-fire regrowth 
-0.1 to +0.1 
Unburned 
0.1 to 0.27 
Low-severity burn 
0.27 to 0.44 
Moderate-low severity burn 
0.44 to 0.66 
Moderate-high severity burn 
> 0.66 
High-severity burn 


| SEVERITY LEVEL  | | dNBR RANGE |
|------------------------------|
| Enhanced Regrowth | |-.50 to  -.10  |
| Unburned       |  |-.10 to .10  |
| Low Severity     | |.10 to .27  |
| Moderate Severity  | | .27 to .66  |
| High Severity     |  | > .660 |


# Band 4 includes wavelengths from 0.76-0.90 µm (NIR) and Band 7 includes wavelengths between 2.09-2.35 µm (SWIR).
B4 - b7 / b4 + b7

```{r create & apply mask}
# calculate modis NBR
modis_nbr <- overlay(all_modis_bands_st_mask[[2]], all_modis_bands_st_mask[[7]],
                     fun=get_veg_index)

# create classification matrix
reclass <- c(cellStats(modis_nbr, min), -.1, 1,
             -.1, .1, 2,
             .1, .27, 3,
             .27, .66, 4,
             .66, cellStats(modis_nbr, max), 5)
# reshape the object into a matrix with columns and rows
reclass_m <- matrix(reclass,
                ncol=3,
                byrow=TRUE)

modis_nbr_cl <- reclassify(modis_nbr,
                     reclass_m)
# reclass data
plot(modis_nbr_cl)

# get summary counts of each class in raster
freq(modis_nbr_cl, useNA='no')
```



