---
layout: single
title: "Work with Precipitation Data in R - 2013 Colorado Floods"
excerpt: "This lesson provides students with an example of a data driven report to emphasize the importance of connecting data, documentation and results."
authors: ['Leah Wasser', 'NEON Data Skills']
modified: '`r format(Sys.time(), "%Y-%m-%d")`'
category: [course-materials]
class-lesson: ['co-floods-1-intro']
permalink: /course-materials/earth-analytics/week-1/co-floods-data-example-r/
nav-title: 'CO Floods Data Example'
week: 1
sidebar:
  nav:
author_profile: false
comments: true
course: "earth-analytics"
order: 2
topics:
  earth-science: ['flood-erosion']
  time-series:
---

{% include toc title="In This Lesson" icon="file-text" %}

Several factors contributed to extreme flooding that occurred in Boulder,
Colorado in 2013. In this lesson we will check our a report that provides some
information about the event.

<div class='notice--success' markdown='1'>

## <i class="fa fa-graduation-cap" aria-hidden="true"></i> Learning Objectives

After completing this tutorial, you will be able to:

* List some of the components of a project that make it more easily re-usable (reproducible) to you when working with other people


## <i class="fa fa-check-square-o fa-2" aria-hidden="true"></i> What you need

You will need a computer with internet access to complete this activity.

</div>

## A data report

Your colleague put together the very informative data report below. The topic of
the report is the 2013 Colorado floods. Examine the report. Then answer the questions
below.


1. What sources of data were used to create the plots?
2. How were the data processed?
3. How did your colleague generate this report? When was it last updated?
4. Who contributed to this report?
5. You'd like to make some changes to the report - can you do that easily? If you
wanted to make changes, what process and tools would you use to make those changes?
6. What units are the precipitation data in?
7. Create a list of the things that would make editing this report easier.


***

## My Report - 2013 Colorado Flood Data

Precipitation Data

A lot of rain impacted Colorado. See below.

```{r load-libraries, echo=FALSE, message=FALSE, results=FALSE, warning=FALSE}
# set your working directory
# setwd("working-dir-path-here")

# load packages
library(ggplot2) # create efficient, professional plots


# set strings as factors to false for everything!
options(stringsAsFactors = FALSE)

#download.file(url = "https://ndownloader.figshare.com/files/7270970",
#              "data/week1/805325-precip_daily_2003-2013.csv")

# import precip data into R data.frame
precip_boulder <- read.csv("data/week1/805325-precip_daily_2003-2013.csv",
                           header = TRUE)


# convert to date/time and retain as a new field
precip_boulder$DateTime <- as.POSIXct(precip_boulder$DATE,
                                  format="%Y%m%d %H:%M")

# assign NoData values to NA
precip_boulder$HPCP[precip_boulder$HPCP==999.99] <- NA


```

```{r daily-summaries, echo=FALSE, message=FALSE, results=FALSE, warning=FALSE, fig.cap="plot 1" }

# convert DATE to a Date class
# (this will strip the time, but that is saved in DateTime)
precip_boulder$DATE <- as.Date(precip_boulder$DateTime, # convert to Date class
                                  format="%Y%m%d %H:%M")
                                  #DATE in the format: YearMonthDay Hour:Minute

precPlot_daily1 <- ggplot(data=precip_boulder,  # the data frame
      aes(DATE, HPCP)) +   # the variables of interest
      geom_bar(stat="identity") +   # create a bar graph
      xlab("Date") + ylab("Precipitation") +  # label the x & y axes
      ggtitle("Daily Precipitation - Boulder Station\n 2003-2013")  # add a title

precPlot_daily1

```

## Fall 2013 Precipitation

Let's check out the data for a few months.


``` {r subset-data, echo=FALSE, message=FALSE, results=FALSE, fig.cap="plot 2 precip" }

# subset 2 months around flood
precip_boulder_AugOct <- precip_boulder %>%
                        filter(DATE >= as.Date('2013-08-15') & DATE <= as.Date('2013-10-15'))


# create new plot
precPlot_flood2 <- ggplot(data=precip_boulder_AugOct, aes(DATE, HPCP)) +
  geom_bar(stat="identity") +
  xlab("Date") + ylab("Precipitation") +
  ggtitle("Daily Total Precipitation Aug - Oct 2013 for Boulder Creek")

precPlot_flood2

```


``` {r all-boulder-station-data, echo=FALSE, message=FALSE, results=FALSE, warning=FALSE, fig.cap="plot 3 discharge"}

download.file(url = "https://ndownloader.figshare.com/files/7271003",
                "data/week1/805333-precip_daily_1948-2013.csv")

# read in data
prec.boulder.all <- read.csv("data/week1/805333-precip_daily_1948-2013.csv",
                           stringsAsFactors = FALSE,
                           header = TRUE)

# assing NoData values to NA
prec.boulder.all$HPCP[prec.boulder.all$HPCP==999.99] <- NA

# format date/time
prec.boulder.all$DateTime <- as.POSIXct(prec.boulder.all$DATE,
                                  format="%Y%m%d %H:%M")
                                  #Date in the format: YearMonthDay Hour:Minute

# create a year-month variable to aggregate to monthly precip
prec.boulder.all$YearMon  = strftime(prec.boulder.all$DateTime, "%Y/%m")

# aggregate by month
prec.boulder.all_monthly <-aggregate(prec.boulder.all$HPCP,   # data to aggregate
																 by=list(prec.boulder.all$YearMon),  # variable to aggregate by
																 FUN=sum,   # take the sum (total) of the precip
																 na.rm=TRUE)  # if the are NA values ignore them
												# if this is FALSE any NA value will prevent a value be totalled

# rename the columns
names(prec.boulder.all_monthly)[names(prec.boulder.all_monthly)=="Group.1"] <- "DATE"
names(prec.boulder.all_monthly)[names(prec.boulder.all_monthly)=="x"] <- "PRECIP"

# re-format YearMon to a Date so x-axis looks good
prec.boulder.all_monthly$DATE <- paste(prec.boulder.all_monthly$DATE,"/01",sep="")
prec.boulder.all_monthly$DATE <- as.Date(prec.boulder.all_monthly$DATE)

# plot data
precPlot_all <- ggplot(data=prec.boulder.all_monthly, aes(DATE,PRECIP)) +
	geom_bar(stat="identity") +
  xlab("Date") + ylab("Precipitation (units)") +
  ggtitle("Total Monthly Precipitation \n Boulder, CO Station")

precPlot_all


```
