---
layout: single
title: "Add a basemap to a R markdown report using ggmap"
excerpt: "This lesson covers creating a basemap with the ggmap package in R. As a backup it also covers the map package given some ongoing bugs with ggmap!"
authors: ['Leah Wasser']
modified: '`r format(Sys.time(), "%Y-%m-%d")`'
category: [course-materials]
class-lesson: ['hw-lidar-r']
permalink: /course-materials/earth-analytics/week-3/ggmap-basemap/
nav-title: 'Create ggmap basemap'
week: 3
sidebar:
  nav:
author_profile: false
comments: true
order: 2
topics:
  data-exploration-and-analysis: ['data-visualization']
  spatial-data-and-gis:
  reproducible-science-and-programming: ['rmarkdown']
---

{% include toc title="In This Lesson" icon="file-text" %}


<div class='notice--success' markdown="1">

## <i class="fa fa-graduation-cap" aria-hidden="true"></i> Learning Objectives

After completing this tutorial, you will be able to:

* Create a quick basemap using ggmap OR
* Create a quick basemap using the maps package

## <i class="fa fa-check-square-o fa-2" aria-hidden="true"></i> What you need

You need `R` and `RStudio` to complete this tutorial. Also you should have
an `earth-analytics` directory setup on your computer with a `/data`
directory with it.

* install **devtools**: `install.packages('devtools')`
* install **ggmap** from github: `devtools::install_github("dkahle/ggmap")`
* install **maps**: `install.packages('maps')`

* [How to Setup R / RStudio](/course-materials/earth-analytics/week-1/setup-r-rstudio/)
* [Setup your working directory](/course-materials/earth-analytics/week-1/setup-working-directory/)

</div>


```{r ggmap-setup }
# install devtools
#install.packages("devtools")
# install ggmap from dev space
# devtools::install_github("dkahle/ggmap")

library(ggmap)
```

## Create basemap

First, let's create a basemap that shows the location of our stream gage.


```{r create-base-map, fig.cap="ggmap base plot" }
myMap <- get_map(location = "Boulder, Colorado",
          source="google",
          maptype="terrain", crop=FALSE,
          zoom=6)
# plot map
ggmap(myMap)

```

Next, let's add a point to our map representing the location of our actual stream
gage data.

Latitude: 40.051667
Longitude: 105.178333

USGS gage 06730200
40°03'06"	105°10'42"

```{r add-points-to-map, fig.cap="ggmap with location point on it. " }
# add points to your map
# creating a sample data.frame with your lat/lon points
lon <- c(-105.178333)
lat <- c(40.051667)
df <- as.data.frame(cbind(lon,lat))

# create a map with a point location for boulder.
ggmap(myMap) + labs(x = "", y = "") +
  geom_point(data = df, aes(x = lon, y = lat, fill = "red", alpha = 0.2), size = 5, shape = 19) +
  guides(fill=FALSE, alpha=FALSE, size=FALSE)


```

## Alternative - maps package

If you can't install ggmap, you can also create nice basemaps using the maps
package. THe maps package allows you to quickly create basemaps of a study areas.
It utilizes a set of vector based layers including layers that map

* countries across the globe
* the United States and associated counties
* and more

You can use the Maps package, combined with the R base plot functions to add
base layers to your map.

```{r load-maps-package}

#install.packages('maps')
library(maps)
```

## Create a basic map of the United States

```{r create-us-map, fig.cap="vector map of the US"}
map('state')
# add a title to your map
title('Map of the United States')
```

Customize colors.
```{r create-us-map-colors, fig.cap="vector map of the US with colors"}
map('state', col="darkgray", fill=TRUE, border="white")
# add a title to your map
title('Map of the United States')
```

Create a map of Colorado with county boundaries.

```{r create-CO-map-colors, fig.cap="vector map of the CO with colors"}
map('county', regions="Colorado", col="darkgray", fill=TRUE, border="grey80")
map('state', regions="Colorado", col="black", add=T)
# add the x, y location of the stream guage using the points
# notice i used two colors adn sized to may the symbol look a little brighter
points(x=-105.178333, y=40.051667, pch=21, col="violetred4", cex=2)
points(x=-105.178333, y=40.051667, pch=8, col="white", cex=1.3)
# add a title to your map
title('County Map of Colorado\nStream gage location')
```

You can stack several map layers using `add=TRUE`. Notice you can create multi-line
titles using `\n`.

```{r create-map, fig.cap="Create final map"}

map('state', fill=TRUE, col="darkgray", border="white", lwd=1)
map(database = "usa", lwd=1, add=T)
# add the adjacent parts of the US; can't forget my homeland
map("state","colorado", col="springgreen",
    lwd=1, fill=TRUE, add=TRUE)
# add gage location
title("Stream gage location\nBoulder, Colorado")
# add the x, y location of hte stream guage using the points
points(x=-105.178333, y=40.051667, pch=8, col="red", cex=1.3)
```
