---
layout: single
title: "Work with Precipitation Data in R - 2013 Colorado Floods"
excerpt: "This lesson walks through the steps need to download and visualize
precipitation data in R to better understand the drivers and impacts of the 2013
Colorado floods."
authors: ['Leah Wasser', 'NEON Data Skills', 'Mariela Perignon']
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
category: [course-materials]
class-lesson: ['co-floods-2-data-r']
permalink: /course-materials/co-floods-precip-r
nav-title: 'Precip Data R'
sidebar:
  nav:
author_profile: false
comments: false
order: 3
---

Several factors contributed to extreme flooding that occurred in Boulder,
Colorado in 2013. In this tutorial, we will import, manipulate and plot precipitation
data downloaded from the National Weather Service's
Cooperative Observer Program.

<div class="success-warning" markdown="1">

### Learning Objectives

After completing this tutorial, you will be able to:

* Import a text file into R.
* Manipulate data from a text file and plot a quantitative output (precipitation over time)
* Publish & share an interactive plot of the data using Plot.ly.
* Subset data by date (if completing Additional Resources code).
* Set a NoData Value to NA in R (if completing Additional Resources code).

### Things You'll Need To Complete This Lesson

Please be sure you have the most current version of R and, preferably,
RStudio to write your code.

 **R Skill Level:** Intermediate - To succeed in this tutorial, you will need to
have basic knowledge for use of the `R` ssoftware program.

### R Libraries to Install:

* **ggplot2:** `install.packages("ggplot2")`
* **plotly:** `install.packages("plotly")`

#### Data Download

<a href="https://ndownloader.figshare.com/files/6780978"> Download Precipitation Data </a>.


#### Directory Structure

Save the data downloaded above in a directory on your computer called data using 
the following path:

First, create a `data` directory (folder) within
your `Documents` directory. If you downloaded the compressed data file above,
unzip this file and place the `distub-events-co13` folder within the `data`
directory you created. If you are planning to access the data directly as
described in the lesson, create a new directory called `distub-events-co13`
wihin your `data` folder and then within it create another directory called
`precip`. If you choose to save your files elsewhere in your file structure, you
will need to modify the directions in the lesson to set your working directory
accordingly.

</div>



# Work with Precipitation Data

## R Libraries

We will use `ggplot2` to efficiently plot our data and `plotly` to create
interactive plots.

```{r load-libraries}
# set your working directory
# setwd("working-dir-path-here")

# load packages
library(ggplot2) # create efficient, professional plots
library(plotly) # create cool interactive plots

# set strings as factors to false for everything!
options(stringsAsFactors = FALSE)

```

```{r load-libraries-hidden, echo=FALSE, results="hide"}
# this package is only added to get the webpage derived from this code to render
# the plotly graphs.  It is NOT needed for any of the analysis or data
# visualizations.

# install.packages("webshot")
# webshot::install_phantomjs()
# library(webshot) # embed the plotly plots
```

## Import Precipitation Data

We will use the `805325-Preciptation_Daily_2003-2013.csv` file
in this analysis. This dataset is the daily precipitation date from the COOP
station 050843 in Boulder, CO for 1 January 2003 through 31 December 2013.

Since the data format is a .csv, we can use `read.csv` to import the data. After
we import the data, we can take a look at the first few lines using `head()`,
which defaults to the first 6 rows, of the `data.frame`. Finally, we can explore
the R object structure.

```{r import-precip}

# import precip data into R data.frame
precip.boulder <- read.csv("disturb-events-co13/precip/805325-precip_daily_2003-2013.csv",
                           header = TRUE)

# view first 6 lines of the data
head(precip.boulder)

# view structure of data
str(precip.boulder)

```

## About the Data

Viewing the structure of these data, we can see that different types of data are included in
this file.

* **STATION** and **STATION_NAME**: Identification of the COOP station.
* **ELEVATION, LATITUDE** and **LONGITUDE**: The spatial location of the station.
* **DATE**: Gives the date in the format: YYYYMMDD HH:MM. Notice that DATE is
currently class `chr`, meaning the data is interpreted as a character class and
not as a date.
* **HPCP**: The total precipitation given in inches
(since we selected `Standard` for the units), recorded
for the hour ending at the time specified by DATE. Importantly, the metadata
(see below) notes that the value 999.99 indicates missing data. Also important,
hours with no precipitation are not recorded.
* **Measurement Flag**: Indicates if there are any abnormalities with the
measurement of the data. Definitions of each flag can be found in Table 2 of the
documentation.
* **Quality Flag**: Indicates if there are any potential quality problems with
the data. Definitions of each flag can be found in Table 3 of the documentation.

Additional information about the data, known as metadata, is available in the
`PRECIP_HLY_documentation.pdf` file that can be downloaded along with the data.
(Note, as of Sept. 2016, there is a mismatch in the data downloaded and the
documentation. The differences are in the units and missing data value:
inches/999.99 (standard) or millimeters/25399.75 (metric)).

## Clean the Data
Before we can start plotting and working with the data we always need to check
several important factors:

* data class: is R interpreting the data the way we expect it. The function
`str()` is an important tools for this.
* NoData Values: We need to know if our data contains a specific value that
means "data are missing" and if this value has been assigned to NA in R.


### Convert Date-Time
As we've noted, the date field is in a character class. We can convert it to a date/time
class that will allow R to correctly interpret the data and allow us to easily
plot the data. We can convert it to a date/time class using `as.POSIXct()`.

```{r convert-date}

# convert to date/time and retain as a new field
precip.boulder$DateTime <- as.POSIXct(precip.boulder$DATE,
                                  format="%Y%m%d %H:%M")
                                  # date in the format: YearMonthDay Hour:Minute

# double check structure
str(precip.boulder$DateTime)

```

* For more information on date/time classes, see the NEON tutorial
<a href="{{ site.baseurl }}/R/time-series-convert-date-time-class-POSIX/" target="_blank"> *Dealing With Dates & Times in R - as.Date, POSIXct, POSIXlt*</a>.

### NoData Values
We've also learned that missing values, also known as NoData
values, are labelled with the placeholder `999.99`. Do we have any NoData values in our data?

``` {r no-data-values-hist}

# histogram - would allow us to see 999.99 NA values
# or other "weird" values that might be NA if we didn't know the NA value
hist(precip.boulder$HPCP)

```

Looking at the histogram, it looks like we have mostly low values (which makes sense) but a few values
up near 1000 -- likely 999.99. We can assign these entries to be `NA`, the value that
R interprets as no data.

``` {r no-data-values}
# assing NoData values to NA
precip.boulder$HPCP[precip.boulder$HPCP==999.99] <- NA

# check that NA values were added;
# we can do this by finding the sum of how many NA values there are
sum(is.na(precip.boulder))

```

There are 94 NA values in our dataset. This is missing data.

#### Questions:

1. Do we need to worry about the missing data?
1. Could they affect our analyses?

This depends on what questions we are asking.  Here we are looking at
general patterns in the data across 10 years. Since we have just over 3650
days in our entire data set, missing 94 probably won't affect the general trends
we are looking at.

Can you think of a research question where we would need to be concerned about
the missing data?

## Plot Precipitation Data
Now that we've cleaned up the data, we can view it. To do this we will plot using
`ggplot()` from the `ggplot2` package.

``` {r plot-precip-hourly}

# plot the data using ggplot2
precPlot_hourly <- ggplot(data=precip.boulder,  # the data frame
      aes(DateTime, HPCP)) +   # the variables of interest
      geom_bar(stat="identity") +   # create a bar graph
      xlab("Date") + ylab("Precipitation (Inches)") +  # label the x & y axes
      ggtitle("Hourly Precipitation - Boulder Station\n 2003-2013")  # add a title

precPlot_hourly

```

As we can see, plots of hourly date lead to very small numbers and is difficult
to represent all information on a figure. Hint: If you can't see any bars on your
plot you might need to zoom in on it.

Plots and comparison of daily precipitation would be easier to view.

## Plot Daily Precipitation

There are several ways to aggregate the data.

#### Daily Plots
If you only want to view the data plotted by date you need to create a column
with only dates (no time) and then re-plot.

```{r daily-summaries }

# convert DATE to a Date class
# (this will strip the time, but that is saved in DateTime)
precip.boulder$DATE <- as.Date(precip.boulder$DateTime, # convert to Date class
                                  format="%Y%m%d %H:%M")
                                  #DATE in the format: YearMonthDay Hour:Minute

# double check conversion
str(precip.boulder$DATE)

precPlot_daily1 <- ggplot(data=precip.boulder,  # the data frame
      aes(DATE, HPCP)) +   # the variables of interest
      geom_bar(stat="identity") +   # create a bar graph
      xlab("Date") + ylab("Precipitation (Inches)") +  # label the x & y axes
      ggtitle("Daily Precipitation - Boulder Station\n 2003-2013")  # add a title

precPlot_daily1

```

R will automatically combine all data from the same day and plot it as one entry.

#### Daily Plots & Data

If you want to record the combined hourly data for each day, you need to create a new data frame to store the daily data. We can
use the `aggregate()` function to combine all the hourly data into daily data.
We will use the date class DATE field we created in the previous code for this.

``` {r daily-summ}

# aggregate the Precipitation (PRECIP) data by DATE
precip.boulder_daily <-aggregate(precip.boulder$HPCP,   # data to aggregate
	by=list(precip.boulder$DATE),  # variable to aggregate by
	FUN=sum,   # take the sum (total) of the precip
	na.rm=TRUE)  # if the are NA values ignore them
	# if this is FALSE any NA value will prevent a value be totalled

# view the results
head(precip.boulder_daily)

```

So we now have daily data but the column names don't mean anything. We can
give them meaningful names by using the `names()` function. Instead of naming the column of
precipitation values with the original `HPCP`, let's call it `PRECIP`.

``` {r rename-fields}

# rename the columns
names(precip.boulder_daily)[names(precip.boulder_daily)=="Group.1"] <- "DATE"
names(precip.boulder_daily)[names(precip.boulder_daily)=="x"] <- "PRECIP"

# double check rename
head(precip.boulder_daily)
```

Now we can plot the daily data.

``` {r daily-prec-plot}

# plot daily data
precPlot_daily <- ggplot(data=precip.boulder_daily,  # the data frame
      aes(DATE, PRECIP)) +   # the variables of interest
      geom_bar(stat="identity") +   # create a bar graph
      xlab("Date") + ylab("Precipitation (inches)") +  # label the x & y axes
      ggtitle("Daily Precipitation - Boulder Station\n 2003-2013")  # add a title

precPlot_daily
```

Compare this plot to the plot we created using the first method. Are they the same?

<i class="fa fa-star"></i> **R Tip:** This manipulation, or aggregation, of data
can also be done with the package `plyr` using the `summarize()` function.
{: .notice}

## Subset the Data

Instead of looking at the data for the full decade, let's now focus on just the
2 months surrounding the flood on 11-15 September. We'll focus on the window from 15
August to 15 October.

Just like aggregating, we can accomplish this by interacting with the larger plot through the graphical interface or
by creating a subset of the data and protting it separately.

#### Subset Within Plot
To see only a subset of the larger plot, we can simply set limits for the
scale on the x-axis with `scale_x_date()`.

``` {r plot-Aug-Oct-2013}

# First, define the limits -- 2 months around the floods
limits <- as.Date(c("2013-08-15", "2013-10-15"))

# Second, plot the data - Flood Time Period
precPlot_flood <- ggplot(data=precip.boulder_daily,
      aes(DATE, PRECIP)) +
      geom_bar(stat="identity") +
      scale_x_date(limits=limits) +
      xlab("Date") + ylab("Precipitation (Inches)") +
      ggtitle("Precipitation - Boulder Station\n August 15 - October 15, 2013")

precPlot_flood

```

Now we can easily see the dramatic rainfall event in mid-September!

<i class="fa fa-star"></i> **R Tip:** If you are using a date-time class, instead
of just a date class, you need to use `scale_x_datetime()`.
{: .notice}

#### Subset The Data

Now let's create a subset of the data and plot it.

``` {r subset-data}

# subset 2 months around flood
precip.boulder_AugOct <- subset(precip.boulder_daily,
                        DATE >= as.Date('2013-08-15') &
												DATE <= as.Date('2013-10-15'))

# check the first & last dates
min(precip.boulder_AugOct$DATE)
max(precip.boulder_AugOct$DATE)

# create new plot
precPlot_flood2 <- ggplot(data=precip.boulder_AugOct, aes(DATE,PRECIP)) +
  geom_bar(stat="identity") +
  xlab("Date") + ylab("Precipitation (inches)") +
  ggtitle("Daily Total Precipitation Aug - Oct 2013 for Boulder Creek")

precPlot_flood2

```


## Interactive Plots - Plotly

Let's turn our plot into an interactive Plotly plot.

``` {r plotly-prep, eval=FALSE}

# setup your plot.ly credentials; if not already set up
Sys.setenv("plotly_username"="your.user.name.here")
Sys.setenv("plotly_api_key"="your.key.here")

```

```{r plotly-precip-data, eval=F }

#view plotly plot in R
ggplotly(precPlot_flood2)

#publish plotly plot to your plot.ly online account when you are happy with it
plotly_POST(precPlot_flood2)

```

<iframe width="900" height="800" frameborder="0" scrolling="no" src="//plot.ly/~leahawasser/161.embed"></iframe>

<div id="challenge" markdown="1">

## Challenge: Plot Precip for Boulder Station Since 1948

The Boulder precipitation station has been recording data since 1948. Use the
steps above to create a plot of the full record of precipitation at this station (1948 - 2013).
The full dataset takes considerable time to download, so we recommend you use the dataset provided in the compressed file ("805333-precip_daily_1948-2013.csv").

As an added challenge, aggregate the data by month instead of by day.

``` {r all-boulder-station-data, echo=FALSE, results="hide", include=FALSE}

# read in data
prec.boulder.all <- read.csv("disturb-events-co13/precip/805333-precip_daily_1948-2013.csv",
                           stringsAsFactors = FALSE,
                           header = TRUE)

# assing NoData values to NA
prec.boulder.all$HPCP[prec.boulder.all$HPCP==999.99] <- NA

# format date/time
prec.boulder.all$DateTime <- as.POSIXct(prec.boulder.all$DATE,
                                  format="%Y%m%d %H:%M")
                                  #Date in the format: YearMonthDay Hour:Minute

# create a year-month variable to aggregate to monthly precip
prec.boulder.all$YearMon  = strftime(prec.boulder.all$DateTime, "%Y/%m")

# aggregate by month
prec.boulder.all_monthly <-aggregate(prec.boulder.all$HPCP,   # data to aggregate
																 by=list(prec.boulder.all$YearMon),  # variable to aggregate by
																 FUN=sum,   # take the sum (total) of the precip
																 na.rm=TRUE)  # if the are NA values ignore them
												# if this is FALSE any NA value will prevent a value be totalled

# rename the columns
names(prec.boulder.all_monthly)[names(prec.boulder.all_monthly)=="Group.1"] <- "DATE"
names(prec.boulder.all_monthly)[names(prec.boulder.all_monthly)=="x"] <- "PRECIP"

# re-format YearMon to a Date so x-axis looks good
prec.boulder.all_monthly$DATE <- paste(prec.boulder.all_monthly$DATE,"/01",sep="")
prec.boulder.all_monthly$DATE <- as.Date(prec.boulder.all_monthly$DATE)

# plot data
precPlot_all <- ggplot(data=prec.boulder.all_monthly, aes(DATE,PRECIP)) +
	geom_bar(stat="identity") +
  xlab("Date") + ylab("Precipitation (inches)") +
  ggtitle("Total Monthly Precipitation \n Boulder, CO Station")

precPlot_all

# create Plotly plot in R
# ggplotly(precPlot_all)

# publish plotly plot to your plot.ly online account when you are happy with it
# plotly_POST(precPlot_all)

```

</div>


## Additional Resources

### Units & Scale
If you are using a dataset downloaded before 2016, the units were in
**hundredths of an inch**. You might want to
create a new column `PRECIP` that contains the data from `HPCP` converted to
inches.

``` {r inches}

# convert from 100th inch by dividing by 100
precip.boulder$PRECIP<-precip.boulder$HPCP/100

# view & check to make sure conversion occured
head(precip.boulder)

```

#### Question
Compare `HPCP` and `PRECIP`. Did we do the conversion correctly?

***
Return to the
<a href="{{ site.basurl }}/teaching-modules/disturb-events-co13/detailed-lesson"> *Ecological Disturbance Teaching Module* by clicking here</a>.
