---
layout: single
title: "Working with vectors and data types in R"
excerpt: "This tutorial introduces vectors in R. It also introduces subsetting,
and working with NA values."
authors: ['Data Carpentry', 'Leah Wasser']
category: [course-materials]
class-lesson: ['get-to-know-r']
permalink: /course-materials/earth-analytics/work-with-data-types-r
nav-title: 'Vectors in R'
dateCreated: 2016-12-13
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
sidebar:
  nav:
author_profile: false
comments: false
order: 3
---

.

<div class='notice--success' markdown="1">

# Learning Objectives
At the end of this activity, you will be able to:

* Understand the structure of and be able to create a vector object in R.
* Be able to define what a `NA` value is in `R` and how it is used in a vector.

## What You Need

You need R and RStudio to complete this tutorial. Also we recommend have you
have an `earth-analytics` directory setup on your computer with a `/data`
directory with it.

* [How to Setup R / R Studio](/course-materials/setup-r-rstudio)
* [Setup your working directory](/course-materials/setup-working-directory)


</div>


## Vectors and data types

A vector is the most common data structure in `R`. A vector is defined as a
a group of values, which most often are either numbers or characters. You can
assign this list of values to an object or variable, just like you
can for a single value. For example we can create a vector of animal weights:

```{r weight-vector }
weight_g <- c(50, 60, 65, 82)
weight_g

```

A vector can also contain characters:

```{r, purl=FALSE}
animals <- c("mouse", "rat", "dog")
animals
```

There are many functions that allow you to inspect the content of a
vector. `length()` tells you how many elements are in a particular vector:

```{r, purl=FALSE}
length(weight_g)
length(animals)
```

An important feature of a vector, is that all of the elements are the same type of data.
The function `class()` indicates the class (the type of element) of an object:

```{r, purl=FALSE}
class(weight_g)
class(animals)
```

The function `str()` provides an overview of the object and the elements it
contains. It is a really useful function when working with large and complex
objects:

```{r, purl=FALSE}
str(weight_g)
str(animals)
```

You can add elements to your vector by using the `c()` function:

```{r, purl=FALSE}
weight_g <- c(weight_g, 90) # adding at the end of the vector
weight_g <- c(30, weight_g) # adding at the beginning of the vector
weight_g
```

What happens here is that we take the original vector `weight_g`, and we are
adding another item first to the end of the other ones, and then another item at
the beginning. We can do this over and over again to grow a vector, or assemble
a dataset. As we program, this may be useful to add results that we are
collecting or calculating.

We just saw 2 of the 6 **atomic vector** types that R uses: `"character"` and
`"numeric"`. These are the basic building blocks that all R objects are built
from. The other 4 are:

* `"logical"` for `TRUE` and `FALSE` (the boolean data type)
* `"integer"` for integer numbers (e.g., `2L`, the `L` indicates to R that it's an integer)
* `"complex"` to represent complex numbers with real and imaginary parts (e.g.,
  `1+4i`) and that's all we're going to say about them
* `"raw"` that we won't discuss further

Vectors are one of the many **data structures** that R uses. Other important
ones are lists (`list`), matrices (`matrix`), data frames (`data.frame`) and
factors (`factor`).

<div class="notice--warning" markdown="1">

# Challenge

* **Question**: We’ve seen that atomic vectors can be of type character,
  numeric, integer, and logical. But what happens if we try to mix these types in
  a single vector?
<!-- * _Answer_: R implicitly converts them to all be the same type -->

* **Question**: What will happen in each of these examples? (hint: use `class()`
  to check the data type of your objects):

```r
num_char <- c(1, 2, 3, 'a')
num_logical <- c(1, 2, 3, TRUE)
char_logical <- c('a', 'b', 'c', TRUE)
tricky <- c(1, 2, 3, '4')
```
* **Question**: Why do you think it happens?
<!-- * _Answer_: Vectors can be of only one data type. R tries to convert (=coerce)
  the content of this vector to find a "common denominator". -->

* **Question**: Can you draw a diagram that represents the hierarchy of the data
  types?
<!-- * _Answer_: `logical -> numeric -> character <-- logical` -->

</div>



```{r, echo=FALSE, eval=FALSE, purl=TRUE}
## We’ve seen that atomic vectors can be of type character, numeric, integer, and
## logical. But what happens if we try to mix these types in a single
## vector?

## What will happen in each of these examples? (hint: use `class()` to
## check the data type of your object)
num_char <- c(1, 2, 3, 'a')

num_logical <- c(1, 2, 3, TRUE)

char_logical <- c('a', 'b', 'c', TRUE)

tricky <- c(1, 2, 3, '4')

## Why do you think it happens?

## Can you draw a diagram that represents the hierarchy of the data
## types?
```


## Subsetting vectors

If we want to extract one or several values from a vector, we must provide one
or several indices in square brackets. For instance:

```{r, results='show', purl=FALSE}
animals <- c("mouse", "rat", "dog", "cat")
animals[2]
animals[c(3, 2)]
```

<fa fa-star></i>**Data Tip:** R indexes start at 1. Programming languages like
Fortran, MATLAB, and R start
counting at 1, because that's what human beings typically do. Languages in the C
family (including C++, Java, Perl, and Python) count from 0 because that's
simpler for computers to do.
{: .notice }

## Subset Vectors

We can subset vectors too. For instance, if you wanted to select only the
values above 50:

```{r, results='show', purl=FALSE}
weight_g > 50    # will return logicals with TRUE for the indices that meet the condition
## so we can use this to select only the values above 50
weight_g[weight_g > 50]
```

You can combine multiple tests using `&` (both conditions are true, AND) or `|`
(at least one of the conditions is true, OR):

```{r, results='show', purl=FALSE}
weight_g[weight_g < 30 | weight_g > 50]
weight_g[weight_g >= 30 & weight_g == 21]
```

When working with vectors of characters, if you are trying to combine many
conditions it can become tedious to type. The function `%in%` allows you to test
if a value is found in a vector:

```{r, results='show', purl=FALSE}
animals <- c("mouse", "rat", "dog", "cat")
animals[animals == "cat" | animals == "rat"] # returns both rat and cat
animals %in% c("rat", "cat", "dog", "duck")
animals[animals %in% c("rat", "cat", "dog", "duck")]
```

> ### Challenge {.challenge}
>
> * Can you figure out why `"four" > "five"` returns `TRUE`?

```{r, echo=FALSE, purl=TRUE}
# * Can you figure out why `"four" > "five"` returns `TRUE`?
```

<!--
```{r, purl=FALSE}
## Answers
## * When using ">" or "<" on strings, R compares their alphabetical order. Here
##   "four" comes after "five", and therefore is "greater than" it.
```
-->



## Missing data

As R was designed to analyze datasets, it includes the concept of missing data
(which is uncommon in other programming languages). Missing data are represented
in vectors as `NA`.

```{r, purl=FALSE}
planets <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter", "Saturn", "Uranus",
             "Neptune", NA)
```

The default setting for most base functions that read data into R is to
interpret `NA` as a missing value. If there are multiple types of missing values
in your dataset, you can extend what R considers a missing value when it reads
the file in.  To do this you supply additional values to the "`na.strings`"
argument of the command to read in the data. For instance, if you wanted to read
in a CSV file named `planets.csv` that had missing values represented as empty
cell, a single blank space, and the value -999, you would use:

```{r, eval=FALSE, purl=FALSE}
planets_df <- read.csv(file = "planets.csv", na.strings = c("", " ", "-999"))
```

When doing operations on numbers, most functions will return `NA` if the data
you are working with include missing values. It is a safer behavior as otherwise
you may overlook that you are dealing with missing data. You can add the
argument `na.rm=TRUE` to calculate the result while ignoring the missing values.

```{r, purl=FALSE}
heights <- c(2, 4, 4, NA, 6)
mean(heights)
max(heights)
mean(heights, na.rm = TRUE)
max(heights, na.rm = TRUE)
```

If your data include missing values, you may want to become familiar with the
functions `is.na()`, `na.omit()`, and `complete.cases()`. See below for
examples.


```{r, purl=FALSE}
## Extract those elements which are not missing values.
heights[!is.na(heights)]

## Returns the object with incomplete cases removed. The returned object is atomic.
na.omit(heights)

## Extract those elements which are complete cases.
heights[complete.cases(heights)]
```

<div class="notice--warning" markdown="1">

# Challenge

* **Question**: Why does the following piece of code give a warning?
```{r, purl=FALSE}
sample <- c(2, 4, 4, "NA", 6)
mean(sample, na.rm = TRUE)
```
<!-- * _Answer_: Because R recognizes the NA in quotes as a character. -->

* **Question**: Why does the warning message say the argument is not numeric?
<!-- * _Answer_: R converts the entire vector to character because of the "NA", and doesn't recognize it as numeric. -->

</div>
