---
layout: single
title: "Objects in R"
excerpt: "This tutorial introduces the R scientific programming language. It is
designed for someone who has not used R before. We will work with precipitation and
stream discharge data for Boulder County."
authors: ['Leah Wasser', 'Data Carpentry']
category: [course-materials]
class-lesson: ['get-to-know-r']
permalink: /course-materials/earth-analytics/objects-in-r
nav-title: 'Objects in R'
dateCreated: 2016-12-13
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
sidebar:
  nav:
author_profile: false
comments: false
order: 2
---

.

<div class='notice--success' markdown="1">

# Learning Objectives
At the end of this activity, you will be able to:

* Be able to create, modify and use objects or variables in `R`.

## What You Need

You need R and RStudio to complete this tutorial. Also we recommend have you
have an `earth-analytics` directory setup on your computer with a `/data`
directory with it.

* [How to Setup R / R Studio](/course-materials/setup-r-rstudio)
* [Setup your working directory](/course-materials/setup-working-directory)

</div>


## Creating objects

You can get output from `R` simply by typing in math in the console

```{r basic-math }
# add 3 + 5
3 + 5
# divide 12 by 7
12/7

```

However, to do useful and interesting things, we need to assign _values_ to
_objects_. To create an object, we need to give it a name followed by the
assignment operator `<-`, and the value we want to give it:

```{r assign-object }
# assign weight_kg to the value of 55
weight_kg <- 55

# view object value
weight_kg

```

## Use Useful Object Names
Objects can be given any name such as `x`, `current_temperature`, or
`subject_id`. However, it is best to use clear and descriptive words when Naming
objects to ensure your code is easy to follow:

1. **Keep object names short:** this makes them easier to read when scanning through code.
2. **Use meaningful names: **For example: `precip` is a more useful name that tells us something about the object compared to `x`
3. **Don't start names with numbers!** Objects that start with a number are NOT VALID in R.
4. **Avoid names that are existing functions in R:** e.g.,
`if`, `else`, `for`, see
[here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Reserved.html)

A few other notes about object names in R:

* R is case sensitive (e.g., `weight_kg` is different from `Weight_kg`).
* Avoid other function names (e.g., `c`, `T`, `mean`, `data`, `df`, `weights`).
* Use nouns for variable names, and verbs for function names.
* Avoid using dots in object names - e.g. `my.dataset` - dots have a special meaning in R (for methods) and other programming languages


## Consistent, Clean Code

Be consistent when you write your code (where you put spaces, how you name
variable, etc.). In R, two popular style guides are
[Hadley Wickham's](http://adv-r.had.co.nz/Style.html) and
[Google's](https://google-styleguide.googlecode.com/svn/trunk/Rguide.xml). Pick
one and use it!


## View object value
When assigning a value to an object, R does not print anything. You can force to
print the value by using parentheses or by typing the name:

```{r, purl=FALSE}
weight_kg <- 55    # doesn't print anything
(weight_kg <- 55)  # but putting parenthesis around the call prints the value of `weight_kg`
weight_kg          # and so does typing the name of the object
```

Now that R has `weight_kg` in memory, we can do arithmetic with it. For
instance, we may want to convert this weight in pounds (weight in pounds is 2.2
times the weight in kg):

```{r, purl=FALSE}
2.2 * weight_kg
```

We can also change a variable's value by assigning it a new one:

```{r, purl=FALSE}
weight_kg <- 57.5
2.2 * weight_kg
```

This means that assigning a value to one variable does not change the values of
other variables.  For example, let's store the animal's weight in pounds in a new
variable, `weight_lb`:

```{r, purl=FALSE}
weight_lb <- 2.2 * weight_kg
```

and then change `weight_kg` to 100.

```{r, purl=FALSE}
weight_kg <- 100
```

What do you think is the current content of the object `weight_lb`? 126.5 or 200?

<div class="notice--warning" markdown="1">

# Challenge

What are the values after each statement in the following?

```{r, purl=FALSE}
mass <- 47.5            # mass?
age  <- 122             # age?
mass <- mass * 2.0      # mass?
age  <- age - 20        # age?
mass_index <- mass/age  # mass_index?
</div>
