---
layout: single
title: "Working with spreadsheet (tabular) data in R"
excerpt: "."
authors: ['Data Carpentry', 'Leah Wasser']
category: [course-materials]
class-lesson: ['get-to-know-r']
permalink: /course-materials/earth-analytics/spreadsheet-data-in-R/
nav-title: 'Spreadsheet Data in R'
dateCreated: 2016-12-13
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
sidebar:
  nav:
author_profile: false
comments: false
order: 5
---

.

<div class='notice--success' markdown="1">

# Learning Objectives
At the end of this activity, you will be able to:

*

## What You Need

You need R and RStudio to complete this tutorial. Also we recommend have you
have an `earth-analytics` directory setup on your computer with a `/data`
directory with it.

* [How to Setup R / R Studio](/course-materials/setup-r-rstudio)
* [Setup your working directory](/course-materials/setup-working-directory)


</div>

In the homework, we used the code below to create a report with knitr.

```{r open-file }

# load the ggplot2 library for plotting
library(ggplot2)

# turn off factors
options(stringsAsFactors = FALSE)

# download data from figshare
# note that we are downloaded the data into your
download.file(url = "https://ndownloader.figshare.com/files/7010681",
              destfile = "data/boulder-precip.csv")
```

First, we used the `download.file` function to download a datafile. In this case,
the data are housed on figshare.com - a popoular data repository that is free
to use if your data are cumulatively smaller than 20gb.

Notice that download.file has two **ARGUMENTS**:

1. url: this is the path to the data file that you wish to download
2. deskfile: this is the location on your computer (in this case: `/data`) and name of the
file when saved (in this case: boulder-precip.csv). So we downloaded a file from
a url on figshare do our data directory. We named that file `boulder-precip.csv`.

Next, we read in the data using the function: `read.csv`.

```{r read-data }
# import data
boulder.precip <- read.csv(file="data/boulder-precip.csv")

```


```{r view-header }
# view first few rows of the data
head(boulder.precip)

# what is the format of the variable in R
str(boulder.precip)
```

## Introduction to the Data.Frame

Data frames are the _de facto_ data structure for most tabular data, and what we
use for statistics and plotting.

A data frame is a collection of vectors of identical lengths. Each vector
represents a column, and each vector can be of a different data type (e.g.,
characters, integers, factors). The `str()` function is useful to inspect the
data types of the columns.

A data frame can be created by hand, but most commonly they are generated by the
functions `read.csv()` or `read.table()`; in other words, when importing
spreadsheets from your hard drive (or the web).

## Extracting / Specifying "columns" By Name

You can extract just one single column from your data.frame using the `$` symbol:

```{r view-column }
# when we download the data we create a dataframe
# view each column of the data frame using it's name (or header)
boulder.precip$DATE

# view the precip column
boulder.precip$PRECIP
```

## Plotting our Data

We can quickly plot our data too.

```{r quick-plot }
# q plot stands for quick plot. Let's use it to plot our data
qplot(x=boulder.precip$DATE,
      y=boulder.precip$PRECIP)

```

<div class="notice--success" markdown="1">

# Challenge

1. List 3 arguments that are available in the `read.csv` function.
2. How do you figure out what working directory you are in?
3. List 2 ways to set the working directory in RStudio
4. Explain what the $ is used for when working with a data.frame in `R`
5. When you use `read.csv` are you executing a: a) function or b) variable ?
</div>
