---
layout: single
title: "Raster analysis workflow in R."
excerpt: "."
authors: ['Leah Wasser']
modified: '`r format(Sys.time(), "%Y-%m-%d")`'
category: [course-materials]
class-lesson: ['week4-review-r']
permalink: /course-materials/earth-analytics/week-4/raster-analysis-workflow-r/
nav-title: 'Raster analysis in R'
week: 4
sidebar:
  nav:
author_profile: false
comments: true
order: 2
---

{% include toc title="In This Lesson" icon="file-text" %}

<div class='notice--success' markdown="1">

## <i class="fa fa-graduation-cap" aria-hidden="true"></i> Learning Objectives

After completing this tutorial, you will be able to:

*

## <i class="fa fa-check-square-o fa-2" aria-hidden="true"></i> What you need

You will need a computer with internet access to complete this lesson and the data for week 4 of the course.

[<i class="fa fa-download" aria-hidden="true"></i> Download Week 4 Data (~500 MB)](https://ndownloader.figshare.com/files/7525363){:data-proofer-ignore='' .btn }

</div>

We can break our data analysis workflow into several steps as follows:

* **Data Processing:** load and "clean" the data. This may include cropping, dealing with NA values, etc
* **Data Exploration:** understand the range and distribution of values in your data. This may involve plotting histograms scatter plots, etc
* **More Data Processing & Analysis:** This may include the final data processing steps that you determined based upon the data exploration phase.
* **Final data analysis:** The final steps of your analysis - often performed using information gathered in the early data processing / exploration stages of your workflow.
* **Presentation:** Refining your results into a final plot or set of plots that are cleaned up, labeled, etc. 

Please note - working with data is not a linear process. There are no defined 
steps. As you work with data more, you will develop your own workflow and approach.

```{r setup}
# load libraries
library(raster)
library(rgdal)

# set working directory
setwd("~/Documents/earth-analytics")
```

Note: try mapview() is a function that allows you to create interactive maps of 
spatial data using leaflet as a base. 

https://cran.r-project.org/web/packages/mapview/index.html 

```{r load-data}
# load data
pre_dtm <- raster("data/week3/BLDR_LeeHill/pre-flood/lidar/pre_DTM.tif")
pre_dsm <- raster("data/week3/BLDR_LeeHill/pre-flood/lidar/pre_DSM.tif")

post_dtm <- raster("data/week3/BLDR_LeeHill/post-flood/lidar/post_DTM.tif")
post_dsm <- raster("data/week3/BLDR_LeeHill/post-flood/lidar/post_DSM.tif")

# import crop extent
crop_ext <- readOGR("data/week3/BLDR_LeeHill", "clip-extent")

```

Calculate the difference.

```{r diff-data }
# calculate dtm difference
dtm_diff_uncropped <- post_dtm - pre_dtm
plot(dtm_diff_uncropped)
```

Next, crop the data.

```{r crop-data, fig.cap="cropped data"}
# crop the data
dtm_diff <- crop(dtm_diff_uncropped, crop_ext)
plot(dtm_diff,
     main="cropped data")

# get a quick glimpse at some of the values for a particular "row"
# note there are a LOT of values in this raster so this won't print all values.
# below i used the head() function to limit the n umber of values returned to 6.
# that way a lot of numbers don't print out in my final knitr output.
head(getValues(dtm_diff, row = 5))

# view max data values
dtm_diff@data@max
dtm_diff@data@min
```


```{r plot-histogram, fig.cap="initial histogram"}
# plot histogram of data
hist(dtm_diff,
     main="distribution of raster cell values in the data",
     xlab="Height (m)")
```

```{r plot-histogram-xlim, fig.cap="initial histogram w xlim to zoom in"}
hist(dtm_diff,
     xlim=c(-2,2),
     main="histogram \nzoomed in to -2 to 2 on the x axis",
     col="brown")

# see how R is breaking up the data
histinfo <- hist(dtm_diff)
histinfo

# how many breaks does R use in the default histogram
length(histinfo$breaks)

# summarize values in the data
summary(dtm_diff, na.rm=T)

```

## Breaks

Above, we saw that we can see how R breaks up our data to create a histogram.
R, by default, creates 35 bins to plot a histogram of our raster data. We can increase the number of breaks or bins that the hist0gram uses with the
argument:

`breaks=number`

In the example below, I used a very large number - 500 so we can see the bins.

```{r plot-histogram-breaks, fig.cap="initial histogram w xlim to zoom in and breaks" }
# where are most of our values?
hist(dtm_diff,
     xlim=c(-2,2),
     breaks=500,
     main="histogram \nzoomed in to -2-2 on the x axis w more breaks")
```

### Histogram with custom breaks

We can create custom breaks or bins in a histogram too. To do this, we pass the same `breaks`
argument a vector of numbers that represent the range for each bin in our histogram.


```{r plot-histogram-breaks2, fig.cap="histogram w custom breaks"}
# We may want to explore breaks in our histogram before plotting our data
hist(dtm_diff,
     breaks=c(-20, -10, -3, -.5, .5, 3, 10, 50),
     main="Histogram with custom breaks",
     xlab="Height (m)",
     col="springgreen")

```

Finally, let's plot the data using the breaks that we created for our histogram
above. We know that there is a high number of cells with a value between -1 and 1.
So let's consider that when we select the colors for our plot.

```{r plot-data, fig.cap="Plot difference dtm."}
# plot dtm difference with breaks
plot(dtm_diff,
     breaks=c(-20, -10, -3, -1, 1, 3, 10, 50),
     col=terrain.colors(7))
```

## Custom plot colors

Next, let's adjust the colors that we use to plot our raster. to do that we will
create a vector of colors, each or which will represent one of our numeric "bins"
of raster values.

This mimics a classified map - we are still exploring our data!

```{r set-colors}
# how many breaks do we have?
# NOTE: we will have one less break than the length of this vector
length(c(-20,-10,-3,-1, 1, 3, 10, 50))
```

Set number of colors based upon how many breaks or bins we have in our data
above we have 8 numbers in our breaks vector. this translates to 7 bins each or which requires a unique color.

```{r plot-with-unique-colors, fig.cap="Plot difference dtm with custom colors."}
# create a vector of colors - one for each "bin" of raster cells
new_colors <- c("palevioletred4", "palevioletred2", "palevioletred1", "ivory1",
                "seagreen1","seagreen2","seagreen4")

plot(dtm_diff,
     breaks=c(-20, -10, -3, -.5, .5, 3, 10, 50),
     col=new_colors,
     legend=F,
     main="Plot of DTM differences\n custom colors")

# make sure legend plots outside of the plot area
par(xpd=T)
# add the legend to the plot
legend(x=dtm_diff@extent@xmax, y=dtm_diff@extent@ymax, # legend location
       legend=c("-20 to -10", "-10 to -3",
                "-3 to -.5", "-.5 to .5",
                ".5 to 3", "3 to 10", "10 to 50"),
       fill=new_colors,
       bty="n",
       cex=.7)

```

## Crop and replot

We can zoom into a part of the raster manually - by first cropping the data
using a manually created plot extent. Then plotting the newly cropped raster subset.

```{r create-new-extent, fig.cap="cropped dtm subset" }

# new_extent <- drawExtent()
new_extent <- extent(473690, 474155.2, 4434849, 4435204)
new_extent
# crop the raster to a smaller area
dtm_diff_crop <- crop(dtm_diff, new_extent)

# Plot the cropped raster
plot(dtm_diff_crop,
     breaks=c(-20, -10, -3, -1, 1, 3, 10, 50),
     col=new_colors,
     legend=F,
     main="Lidar DTM Difference \n cropped subset")

# grab the upper right hand corner coordinates to place the legend.
legendx <- dtm_diff_crop@extent@xmax
legendy <- dtm_diff_crop@extent@ymax

par(xpd=TRUE)
legend(legendx+100, legendy,
       legend=c("-20 to -10", "-10 to -3",
                "-1 to 1", "1 to 3", "3 to 10", "10 to 50"),
       fill=new_colors,
       bty="n",
       cex=.8)
```

```{r clear-par-settings }
dev.off()
```

## Create a final classified dataset

When we have decided what break points work best for our data, then we may chose
to classify the data.

```{r classify-data }

# -20,-10,-3,-1, 1, 3, 10, 50
# create reclass vector
reclass_vector <- c(-20,-10, -3,
                    -10,-3, -2,
                    -3, -.5, -1,
                    -.5, .5, 0,
                    .5, 3, 1,
                    3, 10, 2,
                    10, 50, 3)

reclass_matrix <- matrix(reclass_vector,
                         ncol=3,
                         byrow = T)

reclass_matrix

```

## Reclassify difference raster

```{r reclassify-raster-diff, fig.cap="final plot" }

diff_dtm_rcl <- reclassify(dtm_diff, reclass_matrix)

plot(diff_dtm_rcl,
     col=new_colors,
     legend=F)
par(xpd=T)
legend(dtm_diff@extent@xmax, dtm_diff@extent@ymax,
       legend=c("-20 to -10", "-10 to -3", "-3 to -.5",
                "-.5 to .5", "1 to 3", "3 to 10", "10 to 50"),
       fill=new_colors,
       bty="n",
       cex=.8)
```

Finally view the final histogram

```{r histogram-of-diff-rcl, fig.cap="histogram of differences"}
hist(diff_dtm_rcl,
     main="Histogram of reclassified data",
     xlab="Height Class")

```

Now let's look at one last thing. What would the distribution look like if
we set all values between -.5 to .5 to NA?

```{r reclass, fig.cap="histogram of final cleaned data"}
# create a new raster object
diff_dtm_rcl_na <- diff_dtm_rcl
# assign values between -.5 and .5 to NA
diff_dtm_rcl_na[diff_dtm_rcl_na >= -.5 & diff_dtm_rcl_na <= .5] <- NA
# view histogram
hist(diff_dtm_rcl_na,
     main="Histogram of data \n values between -.5 and .5 set to NA",
     xlab="Difference Class")

# view summary of data
summary(diff_dtm_rcl_na)
```
